### 2016-04-16 Docs/paper/paper.tex

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs2e/llncs}

\usepackage{amssymb,amsmath,bbm}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
\urldef{\mailjr}\path|juste.raimbault@polytechnique.edu|
\urldef{\mailjk}\path|julien.keutchayan@polytechnique.edu|

\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\newcommand{\noun}[1]{\textsc{#1}}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{A Discrepancy-based Framework to Compare Robustness between Multi-Objective Evaluations}

% a short form should be given in case it is too long for the running head
\titlerunning{Discrepancy-based Framework}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{\noun{Juste Raimbault}$^{1,2}$} %and \noun{Julien Keutchayan}$^{3}$}
%
\authorrunning{Discrepancy-based Framework}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{$^{1}$ UMR CNRS 8504 G{\'e}ographie-Cit{\'e}s, Paris, France\\
$^{2}$ UMR-T IFSTTAR 9403 LVMT, Champs-sur-Marne, France\\
%$^{3}$ D{\'e}partement de math{\'e}matiques et de g{\'e}nie industriel,\\
%Ecole Polytechnique de Montr{\'e}al, Montr{\'e}al, Canada\\\medskip
\mailjr\\
%\mailjk
}

\toctitle{Lecture Notes in Computer Science}
\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
Multi-objective evaluation is a necessary aspect when managing complex systems, as the intrinsic complexity of a system is generally closely linked to the potential number of optimization objectives. In the case of spatial socio-technical systems (territorial and urban systems), the recent example of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where such an evaluation is central to the development of sustainable innovative concepts. However, an evaluation makes no sense without its robustness being given (in the ill-defined sense of reliability). Statistical robustness computation methods are highly dependent of underlying statistical models and a model-independent generic framework is still lacking. We propose a generic formulation of such a framework in the case of integrated aggregated indicators, that allows to define a relative measure of robustness taking into account data structure and indicator values. We implement and apply it to virtual data-driven cases of urban systems for Paris districts. First numerical results show the potentialities of this new method. Furthermore, its relative independence of system type and system model appears as an advantage compared to classical statistical robustness methods.
\keywords{Multi-objective Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
\end{abstract}

%%%%%%%%%%%%%%%%
%% Intro
%%%%%%%%%%%%%%%%
\section{Introduction}

%%%%%%%%%%%%%%%%
\subsection{General Context}

Multi-objective problems are organically linked to the complexity of underlying systems. Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration always leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of physical, biological or social system that present emergence and self-organization properties, where objectives can e.g. be the result of heterogeneous interacting agents (see~\cite{newman2011complex} for a large survey of systems concerned by this approach), multi-objective optimization can be explicitly introduced to study or design the system but is often already implicitly ruling the internal mechanisms of the system. The case of socio-technical Complex Systems is particularly interesting as, following~\cite{haken2003face}, they can be seen as hybrid systems embedding social agents into ``technical artifacts'' (sometimes to an unexpected degree creating what \noun{Picon} describes as \emph{cyborgs}~\cite{picon2013smart}), and thus cumulate propensity to be at the origin of multi-objective issues\footnote{We design by \emph{Multi-Objective Evaluation} all practices including the computation of multiple indicators of a system (it can be multi-objective optimization for system design or multi-objective evaluation of an existing system ; our examples being in the second case but the overall method staying rather generic).}. The new notion of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where sustainability implies contradictory objectives. The example of transportation systems, which conception shifted during the second half of the 20th century from cost-benefit analysis to multi-criteria decision-making, is also typical of such systems~\cite{bavoux2005geographie}. Geographical system are now well studied from such a point of view in particular thanks to the integration of multi-objective frameworks within Geographical Information Systems~\cite{carver1991integrating}. As for the micro-case of eco-districts, meso and macro urban planning and design may be made sustainable through indicators evaluation~\cite{jegou2012evaluation}.

A crucial aspect of an evaluation is a certain notion of its reliability, that we call in a fuzzy way \emph{robustness}. Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework. Statistics naturally include such a notion since the construction and estimation of statistical models give diverse indicators of the consistence of results~\cite{launer2014robustness}. The first example that comes to mind is the application of the law of large numbers to obtain the \emph{p-value} of a model fit, that can be interpreted as a confidence measure of estimates. Besides, confidence intervals and \emph{beta-power} are other important indicators of statistical robustness. Concerning multi-objective optimization, in particular through heuristic algorithms (for example genetic algorithms, or operational research solvers), the notion of robustness of a solution concerns more the stability of the solution on the phase space of the corresponding dynamical system. Recent progresses have been done towards unified formulation of robustness for a multi-objective optimization problem, such as~\cite{deb2006introducing} where robust Pareto-front as defined as solutions that are insensitive to small perturbations. In~\cite{1688537}, the notion of degree of robustness is introduced, formalized as a sort of continuity of other solutions in successive neighborhood of a solution.

However, there still lack a generic method to estimate robustness of an evaluation that would be model-independent, i.e. that would be extracted from data structure and indicators but that would not depend on the method used. Some advantages could be for example an \emph{a priori} estimation of potential robustness of an evaluation and thus to decide if the evaluation is worth doing. We propose here a framework answering this issue. It is data-driven and not model-driven in the sense that robustness estimation does not depend on how indicators are computed, as soon as they respect some assumptions that will be detailed in the following.

%%%%%%%%%%%%%%%%
\subsection{Proposed Approach}

\paragraph{Objectives as Spatial Integrals}

Our approach needs to restrict the scope of application. We assume that objectives can be expressed as spatial integrals, so it should apply to any territorial system and our application cases are urban systems. It is not much restrictive in terms of possible indicators if one uses suitable variables and integrated kernels : in a way analog to the method of geographically weighted regression~\cite{brunsdon1998geographically}, any spatial variable can be integrated against regular kernels of variable size and the result will be a spatial aggregation which sense depends on kernel size. The example we use in the following such as conditional means or sums suit well the assumption. Even an already spatially aggregated indicator can be interpreted as a spatial indicator by using a Dirac distribution on the centroid of the corresponding area.

\paragraph{Linearly Aggregated Objectives}

A second assumption we make is that the multi-objective evaluation is done through linear aggregation of objectives. If $(q_i(\vec{x}))_i$ are values of objectives functions, then weights $(w_i)_i$ are defined in order to build the aggregated decision-making function $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, which value determines then the performance of the solution. It is analog to aggregated utility techniques in economics and many domains apply such a method that can be called \emph{Multi-Criteria Decision Analysis}~\cite{wang2009review}. The subtlety lies in the choice of weights, i.e. the shape of the projection function, and various approaches have been developed to find weights depending on the nature of the problem. Recent work~\cite{dobbie2013robustness} proposed to compare robustness of different aggregation techniques through sensitivity analysis, performed by Monte-Carlo simulations on synthetic data. Distribution of biases where obtained for various techniques and some showed to perform significantly better than others. Robustness assessment still depended on models used in that work.

\bigskip

The rest of the paper is organized as follows : section 2 describes intuitively and mathematically the proposed framework ; section 3 then details implementation, data collection for case studies and numerical results including sensitivity analysis ; section 4 finally discuss limitations and potentialities of the method.

%%%%%%%%%%%%%%%%
%% Framework Description
%%%%%%%%%%%%%%%%
\section{Framework Description}

%%%%%%%%%%%%%%%%
\subsection{Intuitive Description}

We describe now the abstract framework allowing theoretically to compare robustnesses of evaluations of two different urban systems. Our framework is a generalization of an empirical method proposed in~\cite{ecodistrictReport} besides a more general benchmarking study on indicator sense and relevance in a sustainability context. Intuitively, it relies on the same simple empirical base which results from the following statements :
\begin{itemize}
\item Urban systems can be seen from the information available, i.e. raw data describing the system. As a data-driven approach, this raw data is the basis of our framework and robustness will be determined by its structure.
\item From data are computed indicators (objective functions). We assume that a choice of indicators is a will to translate particular aspects of the system, that is to capture a realization of an ``urban fact'' (\emph{fait urbain}) in the sense of \noun{Mangin}~\cite{mangin1999projet} - a sort of stylized fact in terms of processes and mechanisms, having various realizations on spatially distinct systems, depending on each precise context.
\item Given many systems and associated indicators, a common space can be built to compare themselves. In that space, data will represent more or less well real systems, depending e.g. on initial scale, precision of data, missing data. We precisely capture that through the notion of point cloud discrepancy, which is a mathematical tool coming from sampling theory which express how a dataset is well distributed in the space it is embedded in~\cite{dick2010digital}. Coupling discrepancies with appropriate weights depending on indicator importance allows to introduce a relative ratio of robustness between two evaluations.
\end{itemize}

%%%%%%%%%%%%%%%%
\subsection{Indicators}

Let $(S_{i})_{1\leq i\leq N}$ be a finite number of geographically disjoints territorial systems, that we assume described through raw data and intermediate indicators, yielding

\[
S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}
\]

with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$ such that each subspace contain real matrices : $\mathcal{X}_{i,k}=\mathbb{R}^{n_{i,k}^{X}p_{i,k}^{X}}$ (the same holding for $\mathcal{Y}_{i}$). We also define an ontological index function $I_{X}(i,k)$ (resp. $I_{Y}(i,k)$) taking integer values which coincide if and only if the two variables have the same ontology in the sense of~\cite{livet2010}, i.e. they are supposed to represent the same real object. We distinguish ``raw data'' $\mathbf{X}_{i}$ from which indicators are computed via explicit deterministic functions, from ``intermediate indicators'' $\mathbf{Y}_{i}$ that are already integrated and can be e.g. outputs of elaborated models simulating some aspects of the urban system.

We define the partial characteristic space of the ``urban fact'' by

\begin{equation}
\begin{split}
(\mathcal{X},\mathcal{Y}) & \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right)\\
& =  \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
\end{split}
\end{equation}

, with $\mathcal{D}_{\mathcal{X}}=\{\mathcal{X}_{i,k}|I(i,k)\textrm{ distincts},n_{i,k}^{X}\mbox{ maximal}\}$
(the same holding for $\mathcal{Y}_{i}$).

It is indeed the abstract space on which indicators are integrated. The indices $c$ introduced as a definition here correspond to different indicators across all systems. This space is the minimal space common to all systems allowing a common definition for indicators on each.

Let $\mathbf{X}_{i,c}$ be the data canonically projected in the corresponding subspace, well defined for all $i$ and all $c$. We make the key assumption that all indicators are computed by integration against a certain kernel, i.e. that for all $c$, there exists $H_{c}$ space of real-valued functions on $(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})$, such that for all $h\in H_{c}$ :
\begin{enumerate}
\item $h$ is ``enough'' regular (tempered distributions e.g.)
\item $q_c=\int_{(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})}h$ is a function describing the ``urban fact'' (the indicator in itself)
\end{enumerate}

Typical concrete example of kernels can be :

\begin{itemize}
\item A mean of rows of $\mathbf{X}_{i,c}$ is computed with $h(x)=x\cdot f_{i,c}(x)$ where $f_{i,c}$ is the density of the distribution of the assumed underlying variable.
\item A rate of elements respecting a given condition $C$, $h(x)=f_{i,c}(x)\chi_{C(x)}$ 
\item For already aggregated variables $\mathbf{Y}$, a Dirac distribution allows to express them also as a kernel integral. 
\end{itemize}

%%%%%%%%%%%%%%%%
\subsection{Aggregation}

As described in introduction, the multi-objective nature of the problem is tackled through linear aggregation. Although it may be seen as a reductive approach, it becomes an asset for the application of integral approximation theorems as the next step of our method. Let define weights for the linear aggregation. We assume the indicators normalized, i.e. $q_c \in [0,1]$, for a more simple construction of relative weights.

Weights result of two components, one deterministic and one ``subjective'' part permitting more flexibility and the inclusion of judgement in the decision process. For $i,c$ and $h_{c}\in H_{c}$ given, the weight $w_{i,c}$ is a mean of :

\begin{itemize}
\item objective weight, corresponding to relative importance of the indicator
\[w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}\]
where $\hat{q}_{i,c}$ is an estimator of $q_{c}$ for data $\mathbf{X}_{i,c}$ (i.e. the effectively calculated value)
\item subjective weight : various potential methods (notations, importance order, etc.), reviewed in~\cite{wang2009review}, are available to give a normalized subjective weight. In practice, these will be set to zero in our first implementation as we test the framework on virtual cases.
\end{itemize}

%%%%%%%%%%%%%%%%
\subsection{Robustness Estimation}

The scene is now set up to be able to estimate the robustness of the evaluation done through the aggregated function. Therefore, we apply an integral approximation method similar to methods introduced in~\cite{varet2010developpement}, since the integrated form of indicators indeed brings the benefits of such powerful theoretical results.

Let $\mathbf{X}_{i,c}=(\vec{X}_{i,c,l})_{1\leq l\leq n_{i,c}}$ and

\[D_{i,c}=Discp_{\tilde{\mathcal{X}}_{c},\infty}(\mathbf{X}_{i,c})\]

the discrepancy of data points cloud~\cite{niederreiter1972discrepancy}.

With $h\in H_{c}$, we have the upper bound on the integral approximation error

\[
\left\Vert \int h_{c}-\frac{1}{n_{i,c}}\sum_{l}h_{c}(\vec{X}_{i,c,l})\right\Vert \leq K\cdot\left|\left|\left|h_{c}\right|\right|\right|\cdot D_{i,c}
\]

where $K$ is a constant regarding data points and objective function. It directly yields

\[
\left\Vert \int\sum w_{i,c}h_{c}-\frac{1}{n_{i,c}}\sum_{l}w_{i,c}h_{c}(\vec{X}_{i,c,l})\right\Vert \leq K\sum_{c}\left|w_{i,c}\right|\left|\left|\left|h_{c}\right|\right|\right|\cdot D_{i,c}
\]

Assuming the error reasonably realized (``worst case'' scenario for knowledge of the theoretical value of aggregated function), we take this upper bound as an approximation of its magnitude. Furthermore, taking normalized indicators implies $\left|\left|\left|h_c\right|\right|\right| = 1$. We propose then to compare error bounds between two evaluations. They depend only on data distribution (equivalent to \emph{statistical robustness}) and on indicators chosen (sort of \emph{ontological robustness}, i.e. do the indicators have a real sense in the chosen context and do their values make sense), and are a way to combine these two type of robustnesses into a single value.

We thus define a \emph{robustness ratio} to compare the robustness of two evaluations by

\begin{equation}
R_{i,i'}=\frac{\sum_{c}w_{i,c}\cdot D_{i,c}}{\sum_{c}w_{i',c}\cdot D_{i',c}}
\end{equation}

By taking then an order relation on evaluations by comparing the position of the ratio to one, it is obvious that we obtain a complete order on all possible evaluations. This ratio should theoretically allow to compare any evaluation of an urban system. To keep an ontological sense to it, it should be used to compare disjoints sub-systems with a reasonable proportion of indicators in common, or the same sub-system with varying indicators. Note that it provides a way to test the influence of indicators on an evaluation by analyzing the sensitivity if the ratio to their removal. On the contrary, finding a ``minimal'' number of indicators each making the ratio strongly vary should be a way to isolate essential parameters ruling the sub-system.

%%%%%%%%%%%%%%%%
%% Results
%%%%%%%%%%%%%%%%
\section{Results}

We apply our framework to a ``toy-model'', as elaborated indicator definition nor raw data collection was not the primary aim of this work, which was more to theoretically introduce the framework. It is however essential for such a data-driven paradigm to demonstrate through simulation the potentialities of the method, what we do here on semi-synthetic data.

%%%%%%%%%%%%%%%%
\subsection{Data Collection}

We base our virtual case on real geographical data, in particular for \emph{arrondissements} of Paris. We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features. We use the street network and position of buildings within the city of Paris\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}. Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source. An example of visualization of used data is shown in Fig.~\ref{fig:data_ex}.

%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/ex_data}
\caption{Example of a visualization of used data for center area of Paris. We use centroids of buildings which polygons are filled in red, and segments of street network drawn in green. Dataset overall consists of around 200000 building features and 100000 road segments.}
\label{fig:data_ex}
\end{figure}
%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
\subsection{Implementation}

Preprocessing of geographical data is made through QGIS~\cite{qgis2011quantum} for performance reasons. Core implementation of the framework is done in R~\cite{team2000r} for the flexibility of data management and statistical computations. Furthermore, the package \texttt{DiceDesign}~\cite{franco20092} written for numerical experiments and sampling purposes, allows an efficient and direct computation of discrepancies. Last but not least, all source code is openly available on the \texttt{git} repository of the project\footnote{at \url{https://github.com/JusteRaimbault/RobustnessDiscrepancy}} for reproducibility purposes~\cite{ram2013git}.

%%%%%%%%%%%%%%%%
\subsection{Virtual Cases}

We work on each district of Paris (from the 1st to the 20th) as an evaluated urban system. We construct random synthetic data associated to spatial features, so each district has to be evaluated many time to obtain mean statistical behavior of toy indicators and robustness ratios. The indicators chosen need to be computed on residential and street network spatial data. We implement two mean kernels and a conditional mean to show different examples, linked to environmental sustainability and quality of life, that are required to be maximized. Note that these indicators have a real meaning but no particular reason to be aggregated, they are chosen here for the convenience of the toy model and the generation of synthetic data. With $a\in \{1\ldots 20\}$ the number of the district, $A(a)$ corresponding spatial extent, $b\in B$ building coordinates and $s\in S$ street segments, we take
\begin{itemize}
\item Complementary of the average daily distance to work with car per individual, approximated by, with $n_{cars}(b)$ number of cars in the building (randomly generated by associated of cars to a number of building proportional to motorization rate $\alpha_m ~ 0.4$ in Paris), $d_w$ distance to work of individuals (generated from the building to a uniformly generated random point in spatial extent of the dataset), and $d_{max}$ the diameter of Paris area,
\[
\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}
\]

\item Complementary of average car flows within the streets in the district, approximated by, with $\varphi(s)$ relative flow in street segment $s$, generated through the minimum of 1 and a log-normal distribution adjusted to have $95\%$ of mass smaller than 1 what mimics the hierarchical distribution of street use (corresponding to betweenness centrality), and $l(s)$ segment length,
\[
\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}
\]

\item Relative length of pedestrian streets $\bar{p}$, computed through a randomly uniformly generated dummy variable adjusted to have a fixed global proportion of segments that are pedestrian.
\end{itemize}

As synthetic data are stochastic, we run the computation for each district $N=50$ times, what was a reasonable compromise between statistical convergence and time required for computation. Table 1 shows results (mean and standard deviations) of indicator values and robustness ratio computation. Obtained standard deviation confirm that this number of repetitions give consistent results. Indicators obtained through a fixed ratio show small variability what may a limit of this toy approach. However, we obtain the interesting result that a majority of districts give more robust evaluations than 1st district, what was expected because of the size and content of this district : it is indeed a small one with large administrative buildings, what means less spatial elements and thus a less robust evaluation following our definition of the robustness.

%%%%%%%%%%%%%%%%%
%% Results table
%%%%%%%%%%%%%%%%

\begin{table}

\begin{tabular}[10pt]{c|c|c|c|c}
Arrdt & $<\bar{d}_w> \pm \sigma (\bar{d}_w)$ & $<\bar{\varphi}> \pm \sigma (\bar{\varphi})$ & $<\bar{p}> \pm \sigma (\bar{p})$ & $R_{i,1}$ \\[3pt]
\hline
1 th & 0.731655 $\pm$ 0.041099 & 0.917462 $\pm$ 0.026637 & 0.191615 $\pm$ 0.052142 & 1.000000 $\pm$ 0.000000\\[3pt]
\hline
2 th & 0.723225 $\pm$ 0.032539 & 0.844350 $\pm$ 0.036085 & 0.209467 $\pm$ 0.058675 & 1.002098 $\pm$ 0.039972\\[3pt]
\hline
3 th & 0.713716 $\pm$ 0.044789 & 0.797313 $\pm$ 0.057480 & 0.185541 $\pm$ 0.065089 & 0.999341 $\pm$ 0.048825\\[3pt]
\hline
4 th & 0.712394 $\pm$ 0.042897 & 0.861635 $\pm$ 0.030859 & 0.201236 $\pm$ 0.044395 & 0.973045 $\pm$ 0.036993\\[3pt]
\hline
5 th & 0.715557 $\pm$ 0.026328 & 0.894675 $\pm$ 0.020730 & 0.209965 $\pm$ 0.050093 & 0.963466 $\pm$ 0.040722\\[3pt]
\hline
6 th & 0.733249 $\pm$ 0.026890 & 0.875613 $\pm$ 0.029169 & 0.206690 $\pm$ 0.054850 & 0.990676 $\pm$ 0.031666\\[3pt]
\hline
7 th & 0.719775 $\pm$ 0.029072 & 0.891861 $\pm$ 0.026695 & 0.209265 $\pm$ 0.041337 & 0.966103 $\pm$ 0.037132\\[3pt]
\hline
8 th & 0.713602 $\pm$ 0.034423 & 0.931776 $\pm$ 0.015356 & 0.208923 $\pm$ 0.036814 & 0.973975 $\pm$ 0.033809\\[3pt]
\hline
9 th & 0.712441 $\pm$ 0.027587 & 0.910817 $\pm$ 0.015915 & 0.202283 $\pm$ 0.049044 & 0.971889 $\pm$ 0.035381\\[3pt]
\hline
10 th & 0.713072 $\pm$ 0.028918 & 0.881710 $\pm$ 0.021668 & 0.210118 $\pm$ 0.040435 & 0.991036 $\pm$ 0.038942\\[3pt]
\hline
11 th & 0.682905 $\pm$ 0.034225 & 0.875217 $\pm$ 0.019678 & 0.203195 $\pm$ 0.047049 & 0.949828 $\pm$ 0.035122\\[3pt]
\hline
12 th & 0.646328 $\pm$ 0.039668 & 0.920086 $\pm$ 0.019238 & 0.198986 $\pm$ 0.023012 & 0.960192 $\pm$ 0.034854\\[3pt]
\hline
13 th & 0.697512 $\pm$ 0.025461 & 0.890253 $\pm$ 0.022778 & 0.201406 $\pm$ 0.030348 & 0.960534 $\pm$ 0.033730\\[3pt]
\hline
14 th & 0.703224 $\pm$ 0.019900 & 0.902898 $\pm$ 0.019830 & 0.205575 $\pm$ 0.038635 & 0.932755 $\pm$ 0.033616\\[3pt]
\hline
15 th & 0.692050 $\pm$ 0.027536 & 0.891654 $\pm$ 0.018239 & 0.200860 $\pm$ 0.024085 & 0.929006 $\pm$ 0.031675\\[3pt]
\hline
16 th & 0.654609 $\pm$ 0.028141 & 0.928181 $\pm$ 0.013477 & 0.202355 $\pm$ 0.017180 & 0.963143 $\pm$ 0.033232\\[3pt]
\hline
17 th & 0.683020 $\pm$ 0.025644 & 0.890392 $\pm$ 0.023586 & 0.198464 $\pm$ 0.033714 & 0.941025 $\pm$ 0.034951\\[3pt]
\hline
18 th & 0.699170 $\pm$ 0.025487 & 0.911382 $\pm$ 0.027290 & 0.188802 $\pm$ 0.036537 & 0.950874 $\pm$ 0.028669\\[3pt]
\hline
19 th & 0.655108 $\pm$ 0.031857 & 0.884214 $\pm$ 0.027816 & 0.209234 $\pm$ 0.032466 & 0.962966 $\pm$ 0.034187\\[3pt]
\hline
20 th & 0.637446 $\pm$ 0.032562 & 0.873755 $\pm$ 0.036792 & 0.196807 $\pm$ 0.026001 & 0.952410 $\pm$ 0.038702\\[3pt]
\hline
\end{tabular}

\bigskip

\caption{Numerical results of simulation for each district with $N=50$ repetitions. Each toy indicator value is given by mean on repetitions and associated standard deviation. Robustness ratio is computed relative to first district (arbitrary choice). A ratio smaller than 1 means that integral bound is smaller for upper district, i.e. that evaluation is more robust for this district. Because of the small size of first district, we expected a majority of district to give ratio smaller than 1, what is confirmed by results, even when adding standard deviations.}

\end{table}

%%%%%%%%%%%%%%%%
%\subsection{Sensitivity Analysis}

% -- NOT POSSIBLE --

%%%%%%%%%%%%%%%%
%% Discussion
%%%%%%%%%%%%%%%%
\section{Discussion}

%%%%%%%%%%%%%%%%
\subsection{Applicability to Real situations}

\paragraph{Integration Within Existing Frameworks}

The applicability of the method on real cases will directly depend on its potential integration within existing framework. Beyond technical difficulties that will surely appear when trying to couple or integrate implementations, more theoretical obstacles could occur, such as fuzzy formulations of functions or data types, consistency issues in databases, etc. Such multi-criteria framework are numerous. Further interesting work would be to attempt integration into an open one, such as e.g. the one described in~\cite{tivadar2014oasis} which calculates various indices of urban segregation.

\paragraph{Availability of Raw Data}

In general, sensitive data such as transportation questionnaires, or very fine granularity census data are not openly available but provided already aggregated at a certain level (for instance French Insee Data are publicly available at administrative city level, smaller units have restricted access). It means that applying the framework may imply complicated data research procedure, its advantage to be flexible being thus reduced through additional constraints.

%%%%%%%%%%%%%%%%
\subsection{Validity of Theoretical Assumptions}

A possible limitation of our approach is the validity of the assumption formulating indicators as spatial integrals. Indeed, many socio-economic indicators are not necessarily depending explicitly on space, and trying to associate them with spatial coordinates may become a slippery slope (e.g. associate individual economic variables with individual residential coordinates will have a sense only if the use of the variable has a relation with space, otherwise it is a non-legitimate artifact). Even indicators which have a spatial value may derive from non-spatial variables, as~\cite{kwan1998space} points out concerning accessibility, when opposing integrated accessibility measures with individual-based non necessarily spatial-based (e.g. individual decisions) measures. Constraining a theoretical representation of a system to fit a framework by changing some of its ontological properties (always in the sense of real meaning of objects) can be understood as a violation of a fundamental rule of modeling and simulation in social science given in~\cite{banos2013HDR}, that is that there can be an universal ``language'' for modeling and some can not express some systems, having for consequence misleading conclusion due to ontology breaking in the case of an overly constrained formulation.

%%%%%%%%%%%%%%%%
\subsection{Framework Generality}

We argue that the fundamental advantage of the proposed framework is its generality and flexibility, since robustness of the evaluations are obtained only through data structure if ones relaxes constraints on the value of weight. Further work should go towards a more general formulation, suppressing for example the linear aggregation assumption. Non-linear aggregation functions would require however to present particular properties regarding integral inequalities. For example, similar results could search in the direction of integral inequalities for Lipschitzian functions such as the one-dimensional results of~\cite{dragomir1999ostrowski}.

%%%%%%%%%%%%%%%%
%% Conclusion
%%%%%%%%%%%%%%%%
\section*{Conclusion}

We have proposed a model-independent framework to compare the robustness of multi-objective evaluations between different urban systems. Based on data discrepancy, it provide a general definition of relative robustness without any assumption on model for the system. Limiting assumption are however the  need of linear aggregation and of indicators that can be expressed through spatial kernel integrals. We propose a toy implementation based on real data for the city of Paris and numerical results confirm general expected behavior. Further work should be oriented towards sensitivity analysis of the method, application to real cases and theoretical assumptions relaxation.

%%%%%%%%%%%%%%%%
%% Biblio
%%%%%%%%%%%%%%%%

\bibliographystyle{unsrt}
\bibliography{biblio/biblio}

\end{document}

### 2016-04-16 Docs/paper/paper.tex
paper_v1

@@
 % first the title is needed
-\title{A Discrepancy-based Framework to Compare Robustness between Multi-Objective Evaluations}
+\title{A Discrepancy-based Framework to Compare Robustness between Multi-Attribute Evaluations}
 
@@
 Multi-objective evaluation is a necessary aspect when managing complex systems, as the intrinsic complexity of a system is generally closely linked to the potential number of optimization objectives.
-In the case of spatial socio-technical systems (territorial and urban systems), the recent example of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where such an evaluation is central to the development of sustainable innovative concepts.
-However, an evaluation makes no sense without its robustness being given (in the ill-defined sense of reliability).
-Statistical robustness computation methods are highly dependent of underlying statistical models and a model-independent generic framework is still lacking.
-We propose a generic formulation of such a framework in the case of integrated aggregated indicators, that allows to define a relative measure of robustness taking into account data structure and indicator values.
-We implement and apply it to virtual data-driven cases of urban systems for Paris districts.
+%In the case of spatial socio-technical systems (territorial and urban systems), the recent example of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where such an evaluation is central to the development of sustainable innovative concepts.
+However, an evaluation makes no sense without its robustness being given (in the sense of its reliability).
+Statistical robustness computation methods are highly dependent of underlying statistical models.
+We propose a formulation of a model-independent framework in the case of integrated aggregated indicators (Multi-attribute evaluation), that allows to define a relative measure of robustness taking into account data structure and indicator values.
+We implement and apply it to a synthetic case of urban systems based on Paris districts geography, and to real data for evaluation of income segregation for Greater Paris metropolitan area.
 First numerical results show the potentialities of this new method.
-Furthermore, its relative independence of system type and system model appears as an advantage compared to classical statistical robustness methods.
-\keywords{Multi-objective Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
+Furthermore, its relative independence to system type and system model may position it as an alternative to classical statistical robustness methods.
+\keywords{Multi-attribute Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
 \end{abstract}
@@
 Multi-objective problems are organically linked to the complexity of underlying systems.
-Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration always leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of physical, biological or social system that present emergence and self-organization properties, where objectives can e.g.
+Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration often leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of non ingeneered  physical, biological or social systems that exhibit emergence and self-organization properties, where objectives can e.g.
 be the result of heterogeneous interacting agents (see~\cite{newman2011complex} for a large survey of systems concerned by this approach), multi-objective optimization can be explicitly introduced to study or design the system but is often already implicitly ruling the internal mechanisms of the system.
@@
 
-A crucial aspect of an evaluation is a certain notion of its reliability, that we call in a fuzzy way \emph{robustness}.
-Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework.
-Statistics naturally include such a notion since the construction and estimation of statistical models give diverse indicators of the consistence of results~\cite{launer2014robustness}.
+A crucial aspect of an evaluation is a certain notion of its reliability, that we call here \emph{robustness}.% Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework.
+Statistics naturally include this notion since the construction and estimation of statistical models give diverse indicators of the consistence of results~\cite{launer2014robustness}.
 The first example that comes to mind is the application of the law of large numbers to obtain the \emph{p-value} of a model fit, that can be interpreted as a confidence measure of estimates.
 Besides, confidence intervals and \emph{beta-power} are other important indicators of statistical robustness.
+Bayesian inference provide also measures of robustness when distribution of parameters are sequentially estimated.
 Concerning multi-objective optimization, in particular through heuristic algorithms (for example genetic algorithms, or operational research solvers), the notion of robustness of a solution concerns more the stability of the solution on the phase space of the corresponding dynamical system.
@@
 
-However, there still lack a generic method to estimate robustness of an evaluation that would be model-independent, i.e.
+However, there still lack generic methods to estimate robustness of an evaluation that would be model-independent, i.e.
 that would be extracted from data structure and indicators but that would not depend on the method used.
 Some advantages could be for example an \emph{a priori} estimation of potential robustness of an evaluation and thus to decide if the evaluation is worth doing.
-We propose here a framework answering this issue.
+We propose here a framework answering this issue in the particular case of Multi-attribute evaluations, i.e.
+when the problem is made unidimensional by objectives aggregation.
 It is data-driven and not model-driven in the sense that robustness estimation does not depend on how indicators are computed, as soon as they respect some assumptions that will be detailed in the following.
@@
 
-Our approach needs to restrict the scope of application.
+%Our approach needs to restrict the scope of application.
 We assume that objectives can be expressed as spatial integrals, so it should apply to any territorial system and our application cases are urban systems.
-It is not much restrictive in terms of possible indicators if one uses suitable variables and integrated kernels : in a way analog to the method of geographically weighted regression~\cite{brunsdon1998geographically}, any spatial variable can be integrated against regular kernels of variable size and the result will be a spatial aggregation which sense depends on kernel size.
+It is not that restrictive in terms of possible indicators if one uses suitable variables and integrated kernels : in a way analog to the method of geographically weighted regression~\cite{brunsdon1998geographically}, any spatial variable can be integrated against regular kernels of variable size and the result will be a spatial aggregation which sense depends on kernel size.
 The example we use in the following such as conditional means or sums suit well the assumption.
@@
 
-A second assumption we make is that the multi-objective evaluation is done through linear aggregation of objectives.
+A second assumption we make is that the multi-objective evaluation is done through linear aggregation of objectives, i.e.
+that we are tackling a multi-attribute optimization problem.
 If $(q_i(\vec{x}))_i$ are values of objectives functions, then weights $(w_i)_i$ are defined in order to build the aggregated decision-making function $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, which value determines then the performance of the solution.
-It is analog to aggregated utility techniques in economics and many domains apply such a method that can be called \emph{Multi-Criteria Decision Analysis}~\cite{wang2009review}.
+It is analog to aggregated utility techniques in economics and is used in many fields.
+%many domains apply such a method that can be called \emph{Multi-Criteria Decision Analysis}~\cite{wang2009review}.
 The subtlety lies in the choice of weights, i.e.
@@
 
-The rest of the paper is organized as follows : section 2 describes intuitively and mathematically the proposed framework ; section 3 then details implementation, data collection for case studies and numerical results including sensitivity analysis ; section 4 finally discuss limitations and potentialities of the method.
+The rest of the paper is organized as follows : section 2 describes intuitively and mathematically the proposed framework ; section 3 then details implementation, data collection for case studies and numerical results for an artificial intra-urban case and a metropolitan real case ; section 4 finally discuss limitations and potentialities of the method.
 
@@
 Our framework is a generalization of an empirical method proposed in~\cite{ecodistrictReport} besides a more general benchmarking study on indicator sense and relevance in a sustainability context.
-Intuitively, it relies on the same simple empirical base which results from the following statements :
+Intuitively, it relies on empirical base resulting from the following axioms :
 \begin{itemize}
@@
 
+Synthesizing these requirements, we propose a notion of \emph{Robustness} of an evaluation that captures both, by combining data reliability with relative importance,
+\begin{enumerate}
+\item \emph{Missing Data} : an evaluation based on more refined datasets will naturally be more robust.
+\item \emph{Indicator importance} : indicators with more relative influence will weight more on the total robustness.
+\end{enumerate}
+
 %%%%%%%%%%%%%%%%
-\subsection{Indicators}
+\subsection{Formal Description}
+
+%%%%%%%%%%%%%%%%
+\paragraph{Indicators}
 
@@
 %%%%%%%%%%%%%%%%
-\subsection{Aggregation}
+\paragraph{Aggregation}
 
-As described in introduction, the multi-objective nature of the problem is tackled through linear aggregation.
+%As described in introduction, the multi-objective nature of the problem is tackled through linear aggregation.
 Although it may be seen as a reductive approach, it becomes an asset for the application of integral approximation theorems as the next step of our method.
+Weighting objectives in multi-attribute decision-making is indeed the crucial point of the processes, and numerous methods are available (see~\cite{wang2009review} for a review for the particular case of sustainable energy management).
 Let define weights for the linear aggregation.
@@
 $q_c \in [0,1]$, for a more simple construction of relative weights.
+For $i,c$ and $h_{c}\in H_{c}$ given, the weight $w_{i,c}$ is simply constituted by the relative importance of the indicator $w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}$ where $\hat{q}_{i,c}$ is an estimator of $q_{c}$ for data $\mathbf{X}_{i,c}$ (i.e.
+the effectively calculated value).
+Note that this step can be extended to any sets of weight attributions, by taking for example $\tilde{w}_{i,c} = w_{i,c} \cdot w'_{i,c}$ if $\mathbf{w}'$ are the weights attributed by the decision-maker.
+We focus here on the relative influence of attributes and thus choose this simple form for weights.
 
-Weights result of two components, one deterministic and one ``subjective'' part permitting more flexibility and the inclusion of judgement in the decision process.
-For $i,c$ and $h_{c}\in H_{c}$ given, the weight $w_{i,c}$ is a mean of :
-
-\begin{itemize}
-\item objective weight, corresponding to relative importance of the indicator
-\[w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}\]
-where $\hat{q}_{i,c}$ is an estimator of $q_{c}$ for data $\mathbf{X}_{i,c}$ (i.e. the effectively calculated value)
-\item subjective weight : various potential methods (notations, importance order, etc.), reviewed in~\cite{wang2009review}, are available to give a normalized subjective weight.
+%  subjective part is shit.
+% subjective weight : various potential methods (notations, importance order, etc.), reviewed in~\cite{wang2009review}, are available to give a normalized subjective weight.
 In practice, these will be set to zero in our first implementation as we test the framework on virtual cases.
-\end{itemize}
 
 %%%%%%%%%%%%%%%%
-\subsection{Robustness Estimation}
+\paragraph{Robustness Estimation}
 
@@
 
+The intuitive sense of this definition is that one compares robustness of evaluations by comparing the highest error done in each based on data structure and relative importance.
+
 By taking then an order relation on evaluations by comparing the position of the ratio to one, it is obvious that we obtain a complete order on all possible evaluations.
@@
 
-We apply our framework to a ``toy-model'', as elaborated indicator definition nor raw data collection was not the primary aim of this work, which was more to theoretically introduce the framework.
+%We apply our framework to a ``toy-model'', as elaborated indicator definition nor raw data collection was not the primary aim of this work, which was more to theoretically introduce the framework.
 It is however essential for such a data-driven paradigm to demonstrate through simulation the potentialities of the method, what we do here on semi-synthetic data.
@@
 %%%%%%%%%%%%%%%%
-\subsection{Data Collection}
-
-We base our virtual case on real geographical data, in particular for \emph{arrondissements} of Paris.
-We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features.
-We use the street network and position of buildings within the city of Paris\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}.
-Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source.
-An example of visualization of used data is shown in Fig.~\ref{fig:data_ex}.
-
-%%%%%%%%%%%%%%%%
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{figures/ex_data}
-\caption{Example of a visualization of used data for center area of Paris.
-We use centroids of buildings which polygons are filled in red, and segments of street network drawn in green.
-Dataset overall consists of around 200000 building features and 100000 road segments.}
-\label{fig:data_ex}
-\end{figure}
-%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%
-\subsection{Implementation}
+\paragraph{Implementation}
 
@@
 %%%%%%%%%%%%%%%%
-\subsection{Virtual Cases}
+\subsection{Implementation on Synthetic Data}
+
+We propose in a first stage to illustrate the implementation with an application to synthetic data and indicators, for intra-urban quality indicators in the city of Paris.
+
+%%%%%%%%%%%%%%%%
+\paragraph{Data Collection}
+
+We base our virtual case on real geographical data, in particular for \emph{arrondissements} of Paris.
+We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features.
+We use the street network and position of buildings within the city of Paris\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}.
+Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source.
+We use centroids of buildings polygons, and segments of street network.
+Dataset overall consists of around $200k$ building features and $100k$ road segments.
+%An example of visualization of used data is shown in Fig.~\ref{fig:data_ex}.
+
+%%%%%%%%%%%%%%%%
+%\begin{figure}
+%\centering
+%\includegraphics[width=\textwidth]{figures/ex_data}
+%\caption{Example of a visualization of used data for center area of Paris.
+We use centroids of buildings which polygons are filled in red, and segments of street network drawn in green.
+Dataset overall consists of around 200000 building features and 100000 road segments.}
+%\label{fig:data_ex}
+%\end{figure}
+%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%
+\paragraph{Virtual Cases}
 
@@
 
-\begin{table}
-
-\begin{tabular}[10pt]{c|c|c|c|c}
+\begin{table}[h!]
+\hspace{-1cm}
+\begin{tabular}[6pt]{c|c|c|c|c}
 Arrdt & $<\bar{d}_w> \pm \sigma (\bar{d}_w)$ & $<\bar{\varphi}> \pm \sigma (\bar{\varphi})$ & $<\bar{p}> \pm \sigma (\bar{p})$ & $R_{i,1}$ \\[3pt]
@@
 %%%%%%%%%%%%%%%%
-%\subsection{Sensitivity Analysis}
+\subsection{Application to a Real Case : Metropolitan Segregation}
 
-% -- NOT POSSIBLE --
+The first example was aimed to show potentialities of the method but was purely synthetic, hence yielding no concrete conclusion nor implications for policy.
+We propose now to apply it to real data for the example of metropolitan segregation.
+
+\paragraph{Data}
+
+We work on income data available for France at an intra-urban level (basic statistical units IRIS) for the year 2011 under the form of summary statistics (deciles if the area is populated enough to ensure anonymity), provided by INSEE\footnote{\texttt{http://www.insee.fr}}.
+Data are associated with geographical extent of statistical units, allowing computation of spatial analysis indicators.
+
+\paragraph{Indicators}
+
+We use here three indicators of segregation integrated on a geographical area.
+Let assume the area divided into covering units $\mathcal{S}_i$ for $1\leq i \leq N$ with centroids $(x_i,y_i)$.
+Each unit has characteristics of population $P_i$ and median income $X_i$.
+We define spatial weights used to quantify strength of geographical interactions between units $i,j$, with $d_{ij}$ euclidian distance between centroids : $w_{ij} = \frac{P_i P_j}{\left(\sum_k P_k\right)^2}\cdot \frac{1}{d_{ij}}$ if $i\neq i$ and $w_{ii} = 0$.
+The normalized indicators are the following
+
+\begin{itemize}
+\item Spatial autocorrelation Moran index, defined as weighted normalized covariance of median income by
+\[
+\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}
+\]
+\item Dissimilarity index (close to Moran but integrating local dissimilarities rather than correlations), with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
+\[
+d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|
+\]
+\item Complementary of the entropy of income distribution that is a way to capture global inequalities
+\[
+\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)
+\]
+\end{itemize}
+
+Numerous measures of segregation with various meanings and at different scales are available, as for example at the level of the unit by comparison of empirical wage distribution with a theoretical null model~\cite{louf2015patterns}.
+The choice here is arbitrary in order to illustrate our method with a reasonable number of dimensions.
+
+%%%%%%%%%%%%%%%%
+\begin{figure}[h!]
+\centering
+%\vspace{-1.5cm}
+\hspace{-3cm}\includegraphics[width=1.2\textwidth]{figures/grandParis_income_moran}
+\vspace{-2.5cm}
+\caption{\textbf{Maps of Metropolitan Segregation.} Maps show yearly median income on basic statistical units (IRIS) for the three departments constituting mainly the Great Paris metropolitan area, and the corresponding local Moran spatial autocorrelation index, defined for unit $i$ as $\rho_i = N/\sum{j}w_{ij} \cdot \frac{\sum_j w_ij (X_j - \bar{X})(X_i - \bar{X})}{\sum_i (X_i - \bar{X})^2}$.
+The most segregated areas coincide with the richest and the poorest, suggesting an increase of segregation in extreme situations.}
+\end{figure}
+%%%%%%%%%%%%%%%%
+
+\paragraph{Results}
+
+We apply our method with these indicators on the Greater Paris area, constituted of four \emph{d{\'e}partements} that are intermediate administrative units.
+The recent creation of a new metropolitan governance system~\cite{gilli2009paris} underlines interrogations on its consistence, and in particular on its relation to intermediate spatial inequalities.
+We show in Fig.
+1 maps of spatial distribution of median income and corresponding local index of autocorrelation.
+We observe the well-known West-East opposition and district disparities inside Paris as they were formulated in various studies, such as~\cite{guerois2009dynamique} through the analysis of real estate transactions dynamics.
+We then apply our framework to answer a concrete question that has implications for urban policy : \textit{how are the evaluation of segregation within different territories sensitive to missing data ?} To do so, we proceed to Monte Carlo simulations (50 repetitions) during which a fixed proportion of data is randomly removed, and the corresponding robustness index is evaluated.
+Simulations are done on each \emph{department} separately, each time relatively to the robustness of the evaluation of full Greater Paris.
+Results are shown in Fig.
+2.
+One area (92) present a slightly better robustness than the reference, what could be explained by its homogeneity and thus highly fiable segregation values.
+Implications for policy that can be drawn are for example direct comparisons between areas : a loss of 8\% of information on Paris area yields an evaluation as robust as with all data in 93 area.
+The latter being a deprived area, the inequality is increased by this relative lower quality of statistical information.
+The study of standard deviations suggest further investigations as different response regimes to data removal seem to exist.
+
+%%%%%%%%%%%%%%%%
+\begin{figure}
+\centering
+%\vspace{-1cm}
+\hspace{-2cm}\includegraphics[width=0.55\textwidth]{figures/alldeps_rob}
+\includegraphics[width=0.55\textwidth]{figures/alldeps_sdrob}
+\caption{\textbf{Sensitivity of robustness to missing data.} \textit{Left.} For each department, Monte Carlo simulations (N=50 repetitions) are used to determine the impact of missing data on robustness of segregation evaluation.
+Robustness ratios are all computed relatively to full metropolitan area with all available data.
+Quasi-linear behavior translates an approximative linear decrease of discrepancy as a function of data size.
+The convergence of richest (75,92) and poorest departments suggest the correction to linear behavior being driven be segregation patterns.
+\textit{Right.} Corresponding standard deviations of robustness ratios.
+Different regimes (in particular 94 against other) unveil phase transitions at different levels of missing data, meaning that the evaluation in 94 is from this point of view more sensitive to missing data.}
+\end{figure}
+%%%%%%%%%%%%%%%%
 
@@
 
+\paragraph{Implications for Decision-making}
+
+The application of our method to concrete decision-making can be thought in different ways.
+First in the case of a comparative multi-attribute decision process, such as the determination of a transportation corridor, the identification of territories on which the evaluation may be flawed (i.e.
+has a poor relative robustness) could allow a more refined focus on these and a corresponding revision of datasets or an adapted revision of weights.
+In any case the overall decision-making process should be made more reliable.
+A second direction lays in the spirit of the real application we have proposed, i.e.
+the sensitivity of evaluation to various parameters such as missing data.
+If a decision appears as reliable because data have few missing points, but the evaluation is very sensitive to it, one will be more careful in the interpretation of results and taking the final decision.
+
 \paragraph{Integration Within Existing Frameworks}
@@
 Further interesting work would be to attempt integration into an open one, such as e.g.
-the one described in~\cite{tivadar2014oasis} which calculates various indices of urban segregation.
+the one described in~\cite{tivadar2014oasis} which calculates various indices of urban segregation, as we have already illustrated the application on metropolitan segregation indexes.
 
@@
 
-In general, sensitive data such as transportation questionnaires, or very fine granularity census data are not openly available but provided already aggregated at a certain level (for instance French Insee Data are publicly available at administrative city level, smaller units have restricted access).
+In general, sensitive data such as transportation questionnaires, or very fine granularity census data are not openly available but provided already aggregated at a certain level (for instance French Insee Data are publicly available at basic statistical unit level or larger areas depending on variables and minimal population constraints, more precise data is under restricted access).
 It means that applying the framework may imply complicated data research procedure, its advantage to be flexible being thus reduced through additional constraints.
@@
 
-We have proposed a model-independent framework to compare the robustness of multi-objective evaluations between different urban systems.
-Based on data discrepancy, it provide a general definition of relative robustness without any assumption on model for the system.
-Limiting assumption are however the  need of linear aggregation and of indicators that can be expressed through spatial kernel integrals.
-We propose a toy implementation based on real data for the city of Paris and numerical results confirm general expected behavior.
-Further work should be oriented towards sensitivity analysis of the method, application to real cases and theoretical assumptions relaxation.
+We have proposed a model-independent framework to compare the robustness of multi-attribute evaluations between different urban systems.
+Based on data discrepancy, it provide a general definition of relative robustness without any assumption on model for the system, but with limiting assumptions that are the need of linear aggregation and of indicators being expressed through spatial kernel integrals.
+We propose a toy implementation based on real data for the city of Paris, numerical results confirming general expected behavior, and an implementation on real data for income segregation on Greater Paris metropolitan areas, giving possible insights into concrete policy questions.
+Further work should be oriented towards sensitivity analysis of the method, application to other real cases and theoretical assumptions relaxation, i.e.
+the relaxation of linear aggregation and spatial integration.
 

### 2016-04-16 Models/functions.R

@@ -103,20 +103,22 @@ robustness <- function(indicators,data){
 }
 
 
-carto <- function(spdf,df,var){
+carto <- function(spdf,df,var,title){
   cols <- carto.pal(pal1 = "green.pal",n1 = 5, pal2 = "red.pal",n2 = 5)
   
-  plot(spdata, border = NA, col = NA, bg = "#E3DEBF")
-  choroLayer(spdf = spdata,
+  #plot(spdata, border = NA, col = NA, bg = "#E3DEBF")
+  choroLayer(spdf = spdf,
              df = df,
              var=var,
              col=cols,
              nclass=10,
              #breaks = quantile(data$NBMEN11,probs=seq(from=0,to=1,by=0.2),na.rm=TRUE),
-             add=TRUE,lwd = 0.1
+             add=FALSE,lwd = 0.1,
+             legend.pos = "topleft",
+             legend.title.txt = title,
+             legend.values.rnd = 2
   )
-  plot(spdata,border = "grey20", lwd=0.1, add=TRUE)
-  
+  plot(spdf,border = "grey20", lwd=0.1, add=TRUE)
 }
 
 

### 2016-04-16 Models/segregAnalysis.R

@@ -7,15 +7,46 @@ source('functions.R')
 
 library(ggplot2)
 
+allres=list()
+for(dep in c("75","92","93","94")){
+  load(paste0('res/missing_ext_',dep,'.RData'))
+  allres[[dep]] = res
+}
 load('res/missing.RData')
+grdParis=res
+ 
+deps = c()
+for(n in names(allres)){deps=append(deps,rep(n,length(allres[[1]])*dim(allres[[1]][[1]])[1]))}
 
-dr = as.tbl(data.frame(rob=unlist(lapply(res,function(d){d[,1]/d[1,1]})),missing=unlist(lapply(res,function(d){d[,2]}))))
-df = dr %>% group_by(missing) %>% summarise(robmean=mean(rob),robsd=sd(rob))
+dr = as.tbl(data.frame(
+  rob=unlist(lapply(allres,function(res){unlist(lapply(res,function(d){d[,1]/grdParis[[1]][1,1]}))})),
+  missing=unlist(lapply(allres,function(res){unlist(lapply(res,function(d){d[,2]}))})),
+  departement=deps
+))
+df = dr %>% group_by(missing,departement) %>% summarise(robmean=mean(rob),robsd=sd(rob))
 
-g = ggplot(data = df,aes(x=missing,y=robmean))
-g+geom_line()+geom_point()+geom_errorbar(aes(ymin=robmean-robsd,ymax=robmean+robsd),width=0.02)
+g = ggplot(data = df,aes(x=missing,y=robmean,colour=departement))
+g+geom_line()+geom_point()+geom_errorbar(aes(ymin=robmean-robsd,ymax=robmean+robsd),width=0.01)+
+   xlab("Missing Data")+ylab("Robustness")
+
+g=ggplot(data=df,aes(x=missing,y=robsd,colour=departement))
+g+geom_line()+geom_smooth()+xlab("Missing Data")+ylab("Std Robustness")
 
 m=lm(robmean~missing,df)
 # -> info in residuals ?
 plot(m$residuals)
+
+
+
+######
+## Mapping
+
+d = loadData('grandParis')
+mo=moran(df = d$jdata,spdf = d$spdata,spatialWeights = d$spweights,var = "medincome",sampling = 1:nrow(d$jdata))
+
+par(mfrow = c(1,2))
+carto(d$spdata,d$jdata,"medincome","Median\nIncome")
+carto(d$spdata,data.frame(d$jdata,moran=mo$localmorans),"moran","Local Moran")
+
+
 

### 2016-07-04 Models/segregation.R
indics

@@ -56,7 +56,7 @@ res <- foreach(i=1:nrep) %dopar% {
   #d = loadData('grandParis')
   d=loadData(paste0('CONTOURS-IRIS',dep))
   jdata=d$jdata;spdata=d$spdata;spweights=d$spweights
-  robs=c();missing=c()
+  robs=c();missing=c();morans=c();entropies=c();dissim=c();discr1=c();discr2=c()
   for(m in seq(from=0.0,to=0.5,by=0.02)){
     sample = sample.int(nrow(jdata),size=floor(nrow(jdata)*(1-m)))
     mo = moran(jdata,spdata,spweights,"medincome",sample)
@@ -66,15 +66,17 @@ res <- foreach(i=1:nrep) %dopar% {
     d2 = discrepancyCriteria(entr$data,type=c('L2'))$DisL2
     robs = append(robs,d1*mo$globalmoran+d1*diss$dissimilarity+d2*entr$entropy)
     missing=append(missing,m)
+    morans=append(morans,mo);entropies=append(entropies,entr);dissim=append(dissim,diss)
+    discr1=append(discr1,d1):discr2=append(discr2,d2)
   }
-  return(data.frame(robs,missing))
+  return(data.frame(robs,missing,morans,entropies,dissim,discr1,discr2))
 }
 
 stopCluster(cl)
 
 show(proc.time()[3]-startTime)
 
-save(res,file=paste0('res/missing_ext_',dep,'.RData'))
+save(res,file=paste0('res/missing_ext_withIndics_',dep,'.RData'))
 
 }
 

### 2016-07-05 Models/segregation.R
rerun_withindics

@@ -42,12 +42,12 @@ departements = c("75","92","93","94","69")
 
 for(dep in departements){
 
-cl <- makeCluster(10)
+cl <- makeCluster(15)
 registerDoParallel(cl)
 
 startTime = proc.time()[3]
 
-nrep = 50
+nrep = 75
 
 #for(k in 1:nrep){
 res <- foreach(i=1:nrep) %dopar% {
@@ -66,8 +66,8 @@ res <- foreach(i=1:nrep) %dopar% {
     d2 = discrepancyCriteria(entr$data,type=c('L2'))$DisL2
     robs = append(robs,d1*mo$globalmoran+d1*diss$dissimilarity+d2*entr$entropy)
     missing=append(missing,m)
-    morans=append(morans,mo);entropies=append(entropies,entr);dissim=append(dissim,diss)
-    discr1=append(discr1,d1):discr2=append(discr2,d2)
+    morans=append(morans,mo$globalmoran);entropies=append(entropies,entr$entropy);dissim=append(dissim,diss$dissimilarity)
+    discr1=append(discr1,d1);discr2=append(discr2,d2)
   }
   return(data.frame(robs,missing,morans,entropies,dissim,discr1,discr2))
 }

### 2016-07-05 Docs/paper/paper.tex
anal

@@
 
-\[D_{i,c}=Discp_{\tilde{\mathcal{X}}_{c},\infty}(\mathbf{X}_{i,c})\]
+\[D_{i,c}=Discp_{\tilde{\mathcal{X}}_{c},L^2}(\mathbf{X}_{i,c})\]
 
-the discrepancy of data points cloud~\cite{niederreiter1972discrepancy}.
+the discrepancy of data points cloud~\cite{niederreiter1972discrepancy}. It is defined as the $L2$-norm of local discrepancy which is for normalized data points $(x_i)$
 
@@
 \vspace{-2.5cm}
-\caption{\textbf{Maps of Metropolitan Segregation.} Maps show yearly median income on basic statistical units (IRIS) for the three departments constituting mainly the Great Paris metropolitan area, and the corresponding local Moran spatial autocorrelation index, defined for unit $i$ as $\rho_i = N/\sum{j}w_{ij} \cdot \frac{\sum_j w_ij (X_j - \bar{X})(X_i - \bar{X})}{\sum_i (X_i - \bar{X})^2}$.
+\caption{\textbf{Maps of Metropolitan Segregation.} Maps show yearly median income on basic statistical units (IRIS) for the three departments constituting mainly the Great Paris metropolitan area, and the corresponding local Moran spatial autocorrelation index, defined for unit $i$ as $\rho_i = N/\sum_{j}w_{ij} \cdot \frac{\sum_{j} w_{ij} (X_j - \bar{X})(X_i - \bar{X})}{\sum_i (X_i - \bar{X})^2}$.
 The most segregated areas coincide with the richest and the poorest, suggesting an increase of segregation in extreme situations.}

### 2016-07-05 Models/segregAnalysis.R

@@ -7,6 +7,16 @@ source('functions.R')
 
 library(ggplot2)
 
+# compte grd Paris with missing = 0
+d = loadData('grandParis')
+jdata=d$jdata;spdata=d$spdata;spweights=d$spweights
+sample = 1:nrow(jdata)
+mo = moran(jdata,spdata,spweights,"medincome",sample)
+diss = dissimilarity(jdata,spdata,spweights,"medincome",sample)
+entr = entropy(jdata,"medincome",sample)
+d1 = discrepancyCriteria(mo$data,type=c('L2'))$DisL2
+d2 = discrepancyCriteria(entr$data,type=c('L2'))$DisL2
+
 allres=list()
 for(dep in c("75","92","93","94")){
   load(paste0('res/missing_ext_',dep,'.RData'))
@@ -17,6 +27,17 @@ grdParis=res
  
 deps = c()
 for(n in names(allres)){deps=append(deps,rep(n,length(allres[[1]])*dim(allres[[1]][[1]])[1]))}
+
+# get max/min of indics
+maxentr = max(c(entr$entropy,sapply(allres,function(res){return(max(res[,4]))})))
+minentr = min(c(entr$entropy,sapply(allres,function(res){return(min(res[,4]))})))
+maxmor = max(c(mo$globalMoran,sapply(allres,function(res){return(max(res[,3]))})))
+minmor = min(c(mo$globalMoran,sapply(allres,function(res){return(min(res[,3]))})))
+maxdiss = max(c(diss$dissimilarity,sapply(allres,function(res){return(max(res[,5]))})))
+mindiss = min(c(diss$dissimilarity,sapply(allres,function(res){return(max(res[,5]))})))
+
+getRob<-function(d){d[6]*(d[3]-minmor)/(maxmor-minmor)+d[6]*(d[5]-mindiss)/(maxdiss-mindiss)+d[7]*(d[4]-minentr)/(maxentr-minentr)}
+
 
 dr = as.tbl(data.frame(
   rob=unlist(lapply(allres,function(res){unlist(lapply(res,function(d){d[,1]/grdParis[[1]][1,1]}))})),

### 2016-07-05 Models/segregation.R

@@ -42,12 +42,12 @@ departements = c("75","92","93","94","69")
 
 for(dep in departements){
 
-cl <- makeCluster(10)
+cl <- makeCluster(2)
 registerDoParallel(cl)
 
 startTime = proc.time()[3]
 
-nrep = 50
+nrep = 2
 
 #for(k in 1:nrep){
 res <- foreach(i=1:nrep) %dopar% {
@@ -67,7 +67,7 @@ res <- foreach(i=1:nrep) %dopar% {
     robs = append(robs,d1*mo$globalmoran+d1*diss$dissimilarity+d2*entr$entropy)
     missing=append(missing,m)
     morans=append(morans,mo);entropies=append(entropies,entr);dissim=append(dissim,diss)
-    discr1=append(discr1,d1):discr2=append(discr2,d2)
+    discr1=append(discr1,d1);discr2=append(discr2,d2)
   }
   return(data.frame(robs,missing,morans,entropies,dissim,discr1,discr2))
 }

### 2016-07-06 Docs/paper/paper.tex
camera_ready_paper

@@
 Statistical robustness computation methods are highly dependent of underlying statistical models.
-We propose a formulation of a model-independent framework in the case of integrated aggregated indicators (Multi-attribute evaluation), that allows to define a relative measure of robustness taking into account data structure and indicator values.
+We propose a formulation of a model-independent framework in the case of integrated aggregated indicators (multi-attribute evaluation), that allows to define a relative measure of robustness taking into account data structure and indicator values.
 We implement and apply it to a synthetic case of urban systems based on Paris districts geography, and to real data for evaluation of income segregation for Greater Paris metropolitan area.
 First numerical results show the potentialities of this new method.
-Furthermore, its relative independence to system type and system model may position it as an alternative to classical statistical robustness methods.
+Furthermore, its relative independence to system type and model may position it as an alternative to classical statistical robustness methods.
 \keywords{Multi-attribute Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
@@
 Multi-objective problems are organically linked to the complexity of underlying systems.
-Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration often leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of non ingeneered  physical, biological or social systems that exhibit emergence and self-organization properties, where objectives can e.g.
+Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration often leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of non engineered physical, biological or social systems that exhibit emergence and self-organization properties, where objectives can e.g.
 be the result of heterogeneous interacting agents (see~\cite{newman2011complex} for a large survey of systems concerned by this approach), multi-objective optimization can be explicitly introduced to study or design the system but is often already implicitly ruling the internal mechanisms of the system.
-The case of socio-technical Complex Systems is particularly interesting as, following~\cite{haken2003face}, they can be seen as hybrid systems embedding social agents into ``technical artifacts'' (sometimes to an unexpected degree creating what \noun{Picon} describes as \emph{cyborgs}~\cite{picon2013smart}), and thus cumulate propensity to be at the origin of multi-objective issues\footnote{We design by \emph{Multi-Objective Evaluation} all practices including the computation of multiple indicators of a system (it can be multi-objective optimization for system design or multi-objective evaluation of an existing system ; our examples being in the second case but the overall method staying rather generic).}.
+The case of socio-technical Complex Systems is particularly interesting as, following~\cite{haken2003face}, they can be seen as hybrid systems embedding social agents into ``technical artifacts'' (sometimes to an unexpected degree creating what \noun{Picon} describes as \emph{cyborgs}~\cite{picon2013smart}), and thus cumulate propensity to be at the origin of multi-objective issues\footnote{We design by \emph{Multi-Objective Evaluation} all practices including the computation of multiple indicators of a system (it can be multi-objective optimization for system design, multi-objective evaluation of an existing system, multi-attribute evaluation ; our particular framework corresponds to the last case).}.
 The new notion of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where sustainability implies contradictory objectives.
@@
 
-A crucial aspect of an evaluation is a certain notion of its reliability, that we call here \emph{robustness}.% Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework.
+A crucial aspect of an evaluation is a certain notion of its reliability, that we call here \emph{robustness}.
+% Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework.
 Statistics naturally include this notion since the construction and estimation of statistical models give diverse indicators of the consistence of results~\cite{launer2014robustness}.
@@
 \item From data are computed indicators (objective functions).
-We assume that a choice of indicators is a will to translate particular aspects of the system, that is to capture a realization of an ``urban fact'' (\emph{fait urbain}) in the sense of \noun{Mangin}~\cite{mangin1999projet} - a sort of stylized fact in terms of processes and mechanisms, having various realizations on spatially distinct systems, depending on each precise context.
-\item Given many systems and associated indicators, a common space can be built to compare themselves.
-In that space, data will represent more or less well real systems, depending e.g.
+We assume that a choice of indicators is an intention to translate particular aspects of the system, i.e.
+to capture a realization of an ``urban fact'' (\emph{fait urbain}) in the sense of \noun{Mangin}~\cite{mangin1999projet} - a sort of stylized fact in terms of processes and mechanisms, having various realizations on spatially distinct systems, depending on each precise context.
+\item Given many systems and associated indicators, a common space can be built to compare them.
+In that space, data represents more or less well real systems, depending e.g.
 on initial scale, precision of data, missing data.
-We precisely capture that through the notion of point cloud discrepancy, which is a mathematical tool coming from sampling theory which express how a dataset is well distributed in the space it is embedded in~\cite{dick2010digital}.
-Coupling discrepancies with appropriate weights depending on indicator importance allows to introduce a relative ratio of robustness between two evaluations.
+We precisely propose to capture that through the notion of point cloud discrepancy, which is a mathematical tool coming from sampling theory expressing how a dataset is distributed in the space it is embedded in~\cite{dick2010digital}.
+%Coupling discrepancies with appropriate weights depending on indicator importance allows to introduce a relative ratio of robustness between two evaluations.
 \end{itemize}
@@
 
-Let $(S_{i})_{1\leq i\leq N}$ be a finite number of geographically disjoints territorial systems, that we assume described through raw data and intermediate indicators, yielding
-
-\[
-S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}
-\]
-
-with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$ such that each subspace contain real matrices : $\mathcal{X}_{i,k}=\mathbb{R}^{n_{i,k}^{X}p_{i,k}^{X}}$ (the same holding for $\mathcal{Y}_{i}$).
+Let $(S_{i})_{1\leq i\leq N}$ be a finite number of geographically disjoints territorial systems, that we assume described through raw data and intermediate indicators, yielding $S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}$ with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$ such that each subspace contain real matrices : $\mathcal{X}_{i,k}=\mathbb{R}^{n_{i,k}^{X}p_{i,k}^{X}}$ (the same holding for $\mathcal{Y}_{i}$).
 We also define an ontological index function $I_{X}(i,k)$ (resp.
@@
 outputs of elaborated models simulating some aspects of the urban system.
-
 We define the partial characteristic space of the ``urban fact'' by
@@
 \begin{equation}
-\begin{split}
-(\mathcal{X},\mathcal{Y}) & \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right)\\
-& =  \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
-\end{split}
+%\begin{split}
+%(\mathcal{X},\mathcal{Y}) & \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right)\\
+%& =  \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
+%\end{split}
+(\mathcal{X},\mathcal{Y}) \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right) = \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
 \end{equation}
 
-, with $\mathcal{D}_{\mathcal{X}}=\{\mathcal{X}_{i,k}|I(i,k)\textrm{ distincts},n_{i,k}^{X}\mbox{ maximal}\}$
+with $\mathcal{D}_{\mathcal{X}}=\{\mathcal{X}_{i,k}|I(i,k)\textrm{ distincts},n_{i,k}^{X}\mbox{ maximal}\}$
 (the same holding for $\mathcal{Y}_{i}$).
-
 It is indeed the abstract space on which indicators are integrated.
@@
 Therefore, we apply an integral approximation method similar to methods introduced in~\cite{varet2010developpement}, since the integrated form of indicators indeed brings the benefits of such powerful theoretical results.
-
-Let $\mathbf{X}_{i,c}=(\vec{X}_{i,c,l})_{1\leq l\leq n_{i,c}}$ and
-
-\[D_{i,c}=Discp_{\tilde{\mathcal{X}}_{c},L^2}(\mathbf{X}_{i,c})\]
-
-the discrepancy of data points cloud~\cite{niederreiter1972discrepancy}. It is defined as the $L2$-norm of local discrepancy which is for normalized data points $(x_i)$
-
+Let $\mathbf{X}_{i,c}=(\vec{X}_{i,c,l})_{1\leq l\leq n_{i,c}}$ and $D_{i,c}=Disc_{\tilde{\mathcal{X}}_{c},L^2}(\mathbf{X}_{i,c})$ the discrepancy of data points cloud\footnote{The discrepancy is defined as the $L2$-norm of local discrepancy which is for normalized data points $\mathbf{X}=(x_{ij})\in \left[0,1\right]^d$, a function of $\mathbf{t}\in \left[0,1\right]^d$ comparing the number of points falling in the corresponding hypercube with its volume, by $disc(\mathbf{t}) = \frac{1}{n}\sum_i \mathbbm{1}_{\prod_j x_{ij}<t_j} - \prod_j t_j$.
+It is a measure of how the point cloud covers the space.}~\cite{niederreiter1972discrepancy}.
 With $h\in H_{c}$, we have the upper bound on the integral approximation error
@@
 
-where $K$ is a constant regarding data points and objective function. It directly yields
+where $K$ is a constant independent of data points and objective function. It directly yields
 
@@
 
-We propose in a first stage to illustrate the implementation with an application to synthetic data and indicators, for intra-urban quality indicators in the city of Paris.
+We propose in a first time to illustrate the implementation with an application to synthetic data and indicators, for intra-urban quality indicators in the city of Paris.
 
@@
 We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features.
-We use the street network and position of buildings within the city of Paris\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}.
+We use the street network and position of buildings within the city of Paris.
+%\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}
 Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source.
@@
 \begin{itemize}
-\item Complementary of the average daily distance to work with car per individual, approximated by, with $n_{cars}(b)$ number of cars in the building (randomly generated by associated of cars to a number of building proportional to motorization rate $\alpha_m ~ 0.4$ in Paris), $d_w$ distance to work of individuals (generated from the building to a uniformly generated random point in spatial extent of the dataset), and $d_{max}$ the diameter of Paris area,
-\[
-\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}
-\]
-
-\item Complementary of average car flows within the streets in the district, approximated by, with $\varphi(s)$ relative flow in street segment $s$, generated through the minimum of 1 and a log-normal distribution adjusted to have $95\%$ of mass smaller than 1 what mimics the hierarchical distribution of street use (corresponding to betweenness centrality), and $l(s)$ segment length,
-\[
-\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}
-\]
+\item Complementary of the average daily distance to work with car per individual, approximated by, with $n_{cars}(b)$ number of cars in the building (randomly generated by associated of cars to a number of building proportional to motorization rate $\alpha_m ~ 0.4$ in Paris), $d_w$ distance to work of individuals (generated from the building to a uniformly generated random point in spatial extent of the dataset), and $d_{max}$ the diameter of Paris area, $\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}$
+\item Complementary of average car flows within the streets in the district, approximated by, with $\varphi(s)$ relative flow in street segment $s$, generated through the minimum of 1 and a log-normal distribution adjusted to have $95\%$ of mass smaller than 1 what mimics the hierarchical distribution of street use (corresponding to betweenness centrality), and $l(s)$ segment length, $\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}$
 
@@
 \end{itemize}
-
-As synthetic data are stochastic, we run the computation for each district $N=50$ times, what was a reasonable compromise between statistical convergence and time required for computation.
-Table 1 shows results (mean and standard deviations) of indicator values and robustness ratio computation.
-Obtained standard deviation confirm that this number of repetitions give consistent results.
-Indicators obtained through a fixed ratio show small variability what may a limit of this toy approach.
-However, we obtain the interesting result that a majority of districts give more robust evaluations than 1st district, what was expected because of the size and content of this district : it is indeed a small one with large administrative buildings, what means less spatial elements and thus a less robust evaluation following our definition of the robustness.
 
@@
 
+As synthetic data are stochastic, we run the computation for each district $N=50$ times, what was a reasonable compromise between statistical convergence and time required for computation.
+Table 1 shows results (mean and standard deviations) of indicator values and robustness ratio computation.
+Obtained standard deviation confirm that this number of repetitions give consistent results.
+Indicators obtained through a fixed ratio show small variability what may a limit of this toy approach.
+However, we obtain the interesting result that a majority of districts give more robust evaluations than 1st district, what was expected because of the size and content of this district : it is indeed a small one with large administrative buildings, what means less spatial elements and thus a less robust evaluation following our definition of the robustness.
+
 %%%%%%%%%%%%%%%%
@@
 \begin{itemize}
-\item Spatial autocorrelation Moran index, defined as weighted normalized covariance of median income by
-\[
-\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}
-\]
-\item Dissimilarity index (close to Moran but integrating local dissimilarities rather than correlations), with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
-\[
-d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|
-\]
-\item Complementary of the entropy of income distribution that is a way to capture global inequalities
-\[
-\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)
-\]
+\item Spatial autocorrelation Moran index, defined as weighted normalized covariance of median income by $\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}$
+\item Dissimilarity index (close to Moran but integrating local dissimilarities rather than correlations), given by $d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|$\\ with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
+\item Complementary of the entropy of income distribution that is a way to capture global inequalities $\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)$
 \end{itemize}
@@
 %\vspace{-1.5cm}
-\hspace{-3cm}\includegraphics[width=1.2\textwidth]{figures/grandParis_income_moran}
+\hspace{-3cm}\includegraphics[width=1.2\textwidth]{figures/grandParis_income_moran}\hspace{-2cm}
 \vspace{-2.5cm}
@@
 We observe the well-known West-East opposition and district disparities inside Paris as they were formulated in various studies, such as~\cite{guerois2009dynamique} through the analysis of real estate transactions dynamics.
-We then apply our framework to answer a concrete question that has implications for urban policy : \textit{how are the evaluation of segregation within different territories sensitive to missing data ?} To do so, we proceed to Monte Carlo simulations (50 repetitions) during which a fixed proportion of data is randomly removed, and the corresponding robustness index is evaluated.
+We then apply our framework to answer a concrete question that has implications for urban policy : \textit{how are the evaluation of segregation within different territories sensitive to missing data ?} To do so, we proceed to Monte Carlo simulations (75 repetitions) during which a fixed proportion of data is randomly removed, and the corresponding robustness index is evaluated with renormalized indicators.
 Simulations are done on each \emph{department} separately, each time relatively to the robustness of the evaluation of full Greater Paris.
@@
 2.
-One area (92) present a slightly better robustness than the reference, what could be explained by its homogeneity and thus highly fiable segregation values.
-Implications for policy that can be drawn are for example direct comparisons between areas : a loss of 8\% of information on Paris area yields an evaluation as robust as with all data in 93 area.
-The latter being a deprived area, the inequality is increased by this relative lower quality of statistical information.
+All areas present a slightly better robustness than the reference, what could be explained by local homogeneity and thus more fiable segregation values.
+Implications for policy that can be drawn are for example direct comparisons between areas : a loss of 30\% of information on 93 area corresponds to a loss of only 25\% in 92 area.
+The first being a deprived area, the inequality is increased by this relative lower quality of statistical information.
 The study of standard deviations suggest further investigations as different response regimes to data removal seem to exist.
@@
 %\vspace{-1cm}
-\hspace{-2cm}\includegraphics[width=0.55\textwidth]{figures/alldeps_rob}
-\includegraphics[width=0.55\textwidth]{figures/alldeps_sdrob}
-\caption{\textbf{Sensitivity of robustness to missing data.} \textit{Left.} For each department, Monte Carlo simulations (N=50 repetitions) are used to determine the impact of missing data on robustness of segregation evaluation.
+\hspace{-2cm}\includegraphics[width=0.55\textwidth]{figures/alldeps_rob_renormindics}
+\includegraphics[width=0.55\textwidth]{figures/alldeps_robsd_renormindics}\hspace{-2cm}
+\caption{\textbf{Sensitivity of robustness to missing data.} \textit{Left.} For each department, Monte Carlo simulations (N=75 repetitions) are used to determine the impact of missing data on robustness of segregation evaluation.
 Robustness ratios are all computed relatively to full metropolitan area with all available data.
 Quasi-linear behavior translates an approximative linear decrease of discrepancy as a function of data size.
-The convergence of richest (75,92) and poorest departments suggest the correction to linear behavior being driven be segregation patterns.
+The similar trajectory of poorest departments (93,94) suggest the correction to linear behavior being driven be segregation patterns.
 \textit{Right.} Corresponding standard deviations of robustness ratios.
-Different regimes (in particular 94 against other) unveil phase transitions at different levels of missing data, meaning that the evaluation in 94 is from this point of view more sensitive to missing data.}
+Different regimes (in particular 93 against others) unveil phase transitions at different levels of missing data, meaning that the evaluation in 94 is from this point of view more sensitive to missing data.}
 \end{figure}
@@
 If a decision appears as reliable because data have few missing points, but the evaluation is very sensitive to it, one will be more careful in the interpretation of results and taking the final decision.
+Further work and testing will however be needed to understand framework behavior in different contexts and be able to pilot its application in various real situations.
 
@@
 individual decisions) measures.
-Constraining a theoretical representation of a system to fit a framework by changing some of its ontological properties (always in the sense of real meaning of objects) can be understood as a violation of a fundamental rule of modeling and simulation in social science given in~\cite{banos2013HDR}, that is that there can be an universal ``language'' for modeling and some can not express some systems, having for consequence misleading conclusion due to ontology breaking in the case of an overly constrained formulation.
+Constraining a theoretical representation of a system to fit a framework by changing some of its ontological properties (always in the sense of real meaning of objects) can be understood as a violation of a fundamental rule of modeling and simulation in social science given in~\cite{banos2013HDR}, that is that there can be an universal ``language'' for modeling and some can not express some systems, having for consequence misleading conclusion due to ontology breaking in the case of an over-constrained formulation.
 
@@
 
+\section*{Acknowledgments}
+
+The author would like to thank Julien Keutchayan (Ecole Polytechnique de Montr{\'e}al) for suggesting the original idea of using discrepancy, and anonymous reviewers for the useful comments and insights.
+
 %%%%%%%%%%%%%%%%

### 2016-07-06 Models/segregAnalysis.R

@@ -19,7 +19,7 @@ d2 = discrepancyCriteria(entr$data,type=c('L2'))$DisL2
 
 allres=list()
 for(dep in c("75","92","93","94")){
-  load(paste0('res/missing_ext_',dep,'.RData'))
+  load(paste0('res/missing_ext_withIndics_',dep,'.RData'))
   allres[[dep]] = res
 }
 load('res/missing.RData')
@@ -29,18 +29,21 @@ deps = c()
 for(n in names(allres)){deps=append(deps,rep(n,length(allres[[1]])*dim(allres[[1]][[1]])[1]))}
 
 # get max/min of indics
-maxentr = max(c(entr$entropy,sapply(allres,function(res){return(max(res[,4]))})))
-minentr = min(c(entr$entropy,sapply(allres,function(res){return(min(res[,4]))})))
-maxmor = max(c(mo$globalMoran,sapply(allres,function(res){return(max(res[,3]))})))
-minmor = min(c(mo$globalMoran,sapply(allres,function(res){return(min(res[,3]))})))
-maxdiss = max(c(diss$dissimilarity,sapply(allres,function(res){return(max(res[,5]))})))
-mindiss = min(c(diss$dissimilarity,sapply(allres,function(res){return(max(res[,5]))})))
+maxentr = max(c(entr$entropy,sapply(allres,function(l){sapply(l,function(res){return(max(res$entropies))})})))
+minentr = min(c(entr$entropy,sapply(allres,function(l){sapply(l,function(res){return(min(res$entropies))})})))
+maxmor = max(c(mo$globalmoran,sapply(allres,function(l){sapply(l,function(res){return(max(res$morans))})})))
+minmor = min(c(mo$globalmoran,sapply(allres,function(l){sapply(l,function(res){return(min(res$morans))})})))
+maxdiss = max(c(diss$dissimilarity,sapply(allres,function(l){sapply(l,function(res){return(max(res$dissim))})})))
+mindiss = min(c(diss$dissimilarity,sapply(allres,function(l){sapply(l,function(res){return(min(res$dissim))})})))
 
 getRob<-function(d){d[6]*(d[3]-minmor)/(maxmor-minmor)+d[6]*(d[5]-mindiss)/(maxdiss-mindiss)+d[7]*(d[4]-minentr)/(maxentr-minentr)}
 
+# grd Paris with all data as reference
[…]

### 2016-08-08 Docs/arxiv/typeinst.tex
proceedings_paper

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs2e/llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Lecture Notes in Computer Science:\\Authors' Instructions
for the Preparation\\of Camera-Ready
Contributions\\to LNCS/LNAI/LNBI Proceedings}

% a short form should be given in case it is too long for the running head
\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Alfred Hofmann%
\thanks{Please note that the LNCS Editorial assumes that all authors have used
the western naming convention, with given names preceding surnames. This determines
the structure of the names in the running heads and the author index.}%
\and Ursula Barth\and Ingrid Haas\and Frank Holzwarth\and\\
Anna Kramer\and Leonie Kunz\and Christine Rei\ss\and\\
Nicole Sator\and Erika Siebert-Cole\and Peter Stra\ss er}
%
\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Springer-Verlag, Computer Science Editorial,\\
Tiergartenstr. 17, 69121 Heidelberg, Germany\\
\mailsa\\
\mailsb\\
\mailsc\\
\url{http://www.springer.com/lncs}}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{Lecture Notes in Computer Science}
\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
The abstract should summarize the contents of the paper and should
contain at least 70 and at most 150 words. It should be written using the
\emph{abstract} environment.
\keywords{We would like to encourage you to list your keywords within
the abstract section}
\end{abstract}

\section{Introduction}

You are strongly encouraged to use \LaTeXe{} for the
preparation of your camera-ready manuscript together with the
corresponding Springer class file \verb+llncs.cls+. Only if you use
\LaTeXe{} can hyperlinks be generated in the online version
of your manuscript.

The \LaTeX{} source of this instruction file for \LaTeX{} users may be
used as a template. This is
located in the ``authors'' subdirectory in
\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
entitled \texttt{typeinst.tex}. There is a separate package for Word 
users. Kindly send the final and checked source
and PDF files of your paper to the Contact Volume Editor. This is
usually one of the organizers of the conference. You should make sure
that the \LaTeX{} and the PDF files are identical and correct and that
only one version of your paper is sent. It is not possible to update
files at a later stage. Please note that we do not need the printed
paper.

We would like to draw your attention to the fact that it is not possible
to modify a paper in any way, once it has been published. This applies
to both the printed book and the online version of the publication.
Every detail, including the order of the names of the authors, should
be checked before the paper is sent to the Volume Editors.

\subsection{Checking the PDF File}

Kindly assure that the Contact Volume Editor is given the name and email
address of the contact author for your paper. The Contact Volume Editor
uses these details to compile a list for our production department at
SPS in India. Once the files have been worked upon, SPS sends a copy of
the final pdf of each paper to its contact author. The contact author is
asked to check through the final pdf to make sure that no errors have
crept in during the transfer or preparation of the files. This should
not be seen as an opportunity to update or copyedit the papers, which is
not possible due to time constraints. Only errors introduced during the
preparation of the files will be corrected.

This round of checking takes place about two weeks after the files have
been sent to the Editorial by the Contact Volume Editor, i.e., roughly
seven weeks before the start of the conference for conference
proceedings, or seven weeks before the volume leaves the printer's, for
post-proceedings. If SPS does not receive a reply from a particular
contact author, within the timeframe given, then it is presumed that the
author has found no errors in the paper. The tight publication schedule
of LNCS does not allow SPS to send reminders or search for alternative
email addresses on the Internet.

In some cases, it is the Contact Volume Editor that checks all the final
pdfs. In such cases, the authors are not involved in the checking phase.

\subsection{Additional Information Required by the Volume Editor}

If you have more than one surname, please make sure that the Volume Editor
knows how you are to be listed in the author index.

\subsection{Copyright Forms}

The copyright form may be downloaded from the ``For Authors"
(Information for LNCS Authors) section of the LNCS Website:
\texttt{www.springer.com/lncs}. Please send your signed copyright form
to the Contact Volume Editor, either as a scanned pdf or by fax or by
courier. One author may sign on behalf of all of the other authors of a
particular paper. Digital signatures are acceptable.

\section{Paper Preparation}

Springer provides you with a complete integrated \LaTeX{} document class
(\texttt{llncs.cls}) for multi-author books such as those in the LNCS
series. Papers not complying with the LNCS style will be reformatted.
This can lead to an increase in the overall number of pages. We would
therefore urge you not to squash your paper.

Please always cancel any superfluous definitions that are
not actually used in your text. If you do not, these may conflict with
the definitions of the macro package, causing changes in the structure
of the text and leading to numerous mistakes in the proofs.

If you wonder what \LaTeX{} is and where it can be obtained, see the
``\textit{LaTeX project site}'' (\url{http://www.latex-project.org})
and especially the webpage ``\textit{How to get it}''
(\url{http://www.latex-project.org/ftp.html}) respectively.

When you use \LaTeX\ together with our document class file,
\texttt{llncs.cls},
your text is typeset automatically in Computer Modern Roman (CM) fonts.
Please do
\emph{not} change the preset fonts. If you have to use fonts other
than the preset fonts, kindly submit these with your files.

Please use the commands \verb+\label+ and \verb+\ref+ for
cross-references and the commands \verb+\bibitem+ and \verb+\cite+ for
references to the bibliography, to enable us to create hyperlinks at
these places.

For preparing your figures electronically and integrating them into
your source file we recommend using the standard \LaTeX{} \verb+graphics+ or
\verb+graphicx+ package. These provide the \verb+\includegraphics+ command.
In general, please refrain from using the \verb+\special+ command.

Remember to submit any further style files and
fonts you have used together with your source files.

\subsubsection{Headings.}

Headings should be capitalized
(i.e., nouns, verbs, and all other words
except articles, prepositions, and conjunctions should be set with an
initial capital) and should,
with the exception of the title, be aligned to the left.
Words joined by a hyphen are subject to a special rule. If the first
word can stand alone, the second word should be capitalized.

Here are some examples of headings: ``Criteria to Disprove
Context-Freeness of Collage Language", ``On Correcting the Intrusion of
Tracing Non-deterministic Programs by Software", ``A User-Friendly and
Extendable Data Distribution System", ``Multi-flip Networks:
Parallelizing GenSAT", ``Self-determinations of Man".

\subsubsection{Lemmas, Propositions, and Theorems.}

The numbers accorded to lemmas, propositions, and theorems, etc. should
appear in consecutive order, starting with Lemma 1, and not, for
example, with Lemma 11.

\subsection{Figures}

For \LaTeX\ users, we recommend using the \emph{graphics} or \emph{graphicx}
package and the \verb+\includegraphics+ command.

Please check that the lines in line drawings are not
interrupted and are of a constant width. Grids and details within the
figures must be clearly legible and may not be written one on top of
the other. Line drawings should have a resolution of at least 800 dpi
(preferably 1200 dpi). The lettering in figures should have a height of
2~mm (10-point type). Figures should be numbered and should have a
caption which should always be positioned \emph{under} the figures, in
contrast to the caption belonging to a table, which should always appear
\emph{above} the table; this is simply achieved as matter of sequence in
your source.

Please center the figures or your tabular material by using the \verb+\centering+
declaration. Short captions are centered by default between the margins
and typeset in 9-point type (Fig.~\ref{fig:example} shows an example).
The distance between text and figure is preset to be about 8~mm, the
distance between figure and caption about 6~mm.

To ensure that the reproduction of your illustrations is of a reasonable
quality, we advise against the use of shading. The contrast should be as
pronounced as possible.

If screenshots are necessary, please make sure that you are happy with
the print quality before you send the files.
\begin{figure}
\centering
\includegraphics[height=6.2cm]{eijkel2}
\caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at
$x_i$ and $x_j$ (\textit{left and right}) lead to the same summed estimate
at $x_s$. This shows a figure consisting of different types of
lines. Elements of the figure described in the caption should be set in
italics, in parentheses, as shown in this sample caption.}
\label{fig:example}
\end{figure}

Please define figures (and tables) as floating objects. Please avoid
using optional location parameters like ``\verb+[h]+" for ``here".

\paragraph{Remark 1.}

In the printed volumes, illustrations are generally black and white
(halftones), and only in exceptional cases, and if the author is
prepared to cover the extra cost for color reproduction, are colored
pictures accepted. Colored pictures are welcome in the electronic
version free of charge. If you send colored figures that are to be
printed in black and white, please make sure that they really are
legible in black and white. Some colors as well as the contrast of
converted colors show up very poorly when printed in black and white.

\subsection{Formulas}

Displayed equations or formulas are centered and set on a separate
line (with an extra line or halfline space above and below). Displayed
expressions should be numbered for reference. The numbers should be
consecutive within each section or within the contribution,
with numbers enclosed in parentheses and set on the right margin --
which is the default if you use the \emph{equation} environment, e.g.,
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt \;  .
\end{equation}

Equations should be punctuated in the same way as ordinary
text but with a small space before the end punctuation mark.

\subsection{Footnotes}

The superscript numeral used to refer to a footnote appears in the text
either directly after the word to be discussed or -- in relation to a
phrase or a sentence -- following the punctuation sign (comma,
semicolon, or period). Footnotes should appear at the bottom of
the
normal text area, with a line of about 2~cm set
immediately above them.\footnote{The footnote numeral is set flush left
and the text follows with the usual word spacing.}

\subsection{Program Code}

Program listings or program commands in the text are normally set in
typewriter font, e.g., CMTT10 or Courier.

\medskip

\noindent
{\it Example of a Computer Program}
\begin{verbatim}
program Inflation (Output)
  {Assuming annual inflation rates of 7%, 8%, and 10%,...
   years};
   const
     MaxYears = 10;
   var
     Year: 0..MaxYears;
     Factor1, Factor2, Factor3: Real;
   begin
     Year := 0;
     Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
     WriteLn('Year  7% 8% 10%'); WriteLn;
     repeat
       Year := Year + 1;
       Factor1 := Factor1 * 1.07;
       Factor2 := Factor2 * 1.08;
       Factor3 := Factor3 * 1.10;
       WriteLn(Year:5,Factor1:7:3,Factor2:7:3,Factor3:7:3)
     until Year = MaxYears
end.
\end{verbatim}
%
\noindent
{\small (Example from Jensen K., Wirth N. (1991) Pascal user manual and
report. Springer, New York)}

\subsection{Citations}

For citations in the text please use
square brackets and consecutive numbers: \cite{jour}, \cite{lncschap},
\cite{proceeding1} -- provided automatically
by \LaTeX 's \verb|\cite| \dots\verb|\bibitem| mechanism.

\subsection{Page Numbering and Running Heads}

There is no need to include page numbers. If your paper title is too
long to serve as a running head, it will be shortened. Your suggestion
as to how to shorten it would be most welcome.

\section{LNCS Online}

The online version of the volume will be available in LNCS Online.
Members of institutes subscribing to the Lecture Notes in Computer
Science series have access to all the pdfs of all the online
publications. Non-subscribers can only read as far as the abstracts. If
they try to go beyond this point, they are automatically asked, whether
they would like to order the pdf, and are given instructions as to how
to do so.

Please note that, if your email address is given in your paper,
it will also be included in the meta data of the online version.

\section{BibTeX Entries}

The correct BibTeX entries for the Lecture Notes in Computer Science
volumes can be found at the following Website shortly after the
publication of the book:
\url{http://www.informatik.uni-trier.de/~ley/db/journals/lncs.html}

\subsubsection*{Acknowledgments.} The heading should be treated as a
subsubsection heading and should not be assigned a number.

\section{The References Section}\label{references}

In order to permit cross referencing within LNCS-Online, and eventually
between different publishers and their online databases, LNCS will,
from now on, be standardizing the format of the references. This new
feature will increase the visibility of publications and facilitate
academic research considerably. Please base your references on the
examples below. References that don't adhere to this style will be
reformatted by Springer. You should therefore check your references
thoroughly when you receive the final pdf of your paper.
The reference section must be complete. You may not omit references.
Instructions as to where to find a fuller version of the references are
not permissible.

We only accept references written using the latin alphabet. If the title
of the book you are referring to is in Russian or Chinese, then please write
(in Russian) or (in Chinese) at the end of the transcript or translation
of the title.

The following section shows a sample reference list with entries for
journal articles \cite{jour}, an LNCS chapter \cite{lncschap}, a book
\cite{book}, proceedings without editors \cite{proceeding1} and
\cite{proceeding2}, as well as a URL \cite{url}.
Please note that proceedings published in LNCS are not cited with their
full titles, but with their acronyms!

\begin{thebibliography}{4}

\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
Subsequences. J. Mol. Biol. 147, 195--197 (1981)

\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
Composing a Complex Biological Workflow through Web Services. In: Nagel,
W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
pp. 1148--1158. Springer, Heidelberg (2006)

\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
Infrastructure. Morgan Kaufmann, San Francisco (1999)

\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
Information Services for Distributed Resource Sharing. In: 10th IEEE
International Symposium on High Performance Distributed Computing, pp.
181--184. IEEE Press, New York (2001)

\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
Grid: an Open Grid Services Architecture for Distributed Systems
Integration. Technical report, Global Grid Forum (2002)

\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}

\end{thebibliography}

\section*{Appendix: Springer-Author Discount}

LNCS authors are entitled to a 33.3\% discount off all Springer
publications. Before placing an order, the author should send an email, 
giving full details of his or her Springer publication,
to \url{orders-HD-individuals@springer.com} to obtain a so-called token. This token is a
number, which must be entered when placing an order via the Internet, in
order to obtain the discount.

\section{Checklist of Items to be Sent to Volume Editors}
Here is a checklist of everything the volume editor requires from you:

\begin{itemize}
\settowidth{\leftmargin}{{\Large$\square$}}\advance\leftmargin\labelsep
\itemsep8pt\relax
\renewcommand\labelitemi{{\lower1.5pt\hbox{\Large$\square$}}}

\item The final \LaTeX{} source files
\item A final PDF file
\item A copyright form, signed by one author on behalf of all of the
authors of the paper.
\item A readme giving the name and email address of the
corresponding author.
\end{itemize}
\end{document}

### 2016-08-08 Docs/paper/paper.tex
proceedings_paper

@@
 
-\documentclass[runningheads,a4paper]{llncs2e/llncs}
+\documentclass[runningheads,a4paper]{llncs}
 
@@
 %%%%%%%%%%%%%%%%
-%\begin{figure}
-%\centering
-%\includegraphics[width=\textwidth]{figures/ex_data}
-%\caption{Example of a visualization of used data for center area of Paris.
-We use centroids of buildings which polygons are filled in red, and segments of street network drawn in green.
-Dataset overall consists of around 200000 building features and 100000 road segments.}
-%\label{fig:data_ex}
-%\end{figure}
-%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%
 \paragraph{Virtual Cases}
@@
 %\vspace{-1.5cm}
-\hspace{-3cm}\includegraphics[width=1.2\textwidth]{figures/grandParis_income_moran}\hspace{-2cm}
+\hspace{-3cm}\includegraphics[width=1.2\textwidth]{grandParis_income_moran.pdf}\hspace{-2cm}
 \vspace{-2.5cm}
@@
 %\vspace{-1cm}
-\hspace{-2cm}\includegraphics[width=0.55\textwidth]{figures/alldeps_rob_renormindics}
-\includegraphics[width=0.55\textwidth]{figures/alldeps_robsd_renormindics}\hspace{-2cm}
+\hspace{-2cm}\includegraphics[width=0.55\textwidth]{alldeps_rob_renormindics.pdf}
+\includegraphics[width=0.55\textwidth]{alldeps_robsd_renormindics.pdf}\hspace{-2cm}
 \caption{\textbf{Sensitivity of robustness to missing data.} \textit{Left.} For each department, Monte Carlo simulations (N=75 repetitions) are used to determine the impact of missing data on robustness of segregation evaluation.
@@
 \bibliographystyle{unsrt}
-\bibliography{biblio/biblio}
+\bibliography{biblio}
 

### 2016-12-09 Docs/Communication/discrepancy.tex
skeleton_slides

\documentclass[english,11pt]{beamer}

\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Proba}{\mathbb{P}}

\newcommand{\Covb}[2]{\ensuremath{\Cov\!\left[#1,#2\right]}}
\newcommand{\Eb}[1]{\ensuremath{\E\!\left[#1\right]}}
\newcommand{\Pb}[1]{\ensuremath{\Proba\!\left[#1\right]}}
\newcommand{\Varb}[1]{\ensuremath{\Var\!\left[#1\right]}}

% norm
\newcommand{\norm}[1]{\| #1 \|}

\newcommand{\indep}{\rotatebox[origin=c]{90}{$\models$}}

\usepackage{mathptmx,amsmath,amssymb,graphicx,bibentry,bbm,babel,ragged2e}

\makeatletter

\newcommand{\noun}[1]{\textsc{#1}}
\newcommand{\jitem}[1]{\item \begin{justify} #1 \end{justify} \vfill{}}
\newcommand{\sframe}[2]{\frame{\frametitle{#1} #2}}

\newenvironment{centercolumns}{\begin{columns}[c]}{\end{columns}}
%\newenvironment{jitem}{\begin{justify}\begin{itemize}}{\end{itemize}\end{justify}}

\usetheme{Warsaw}
\setbeamertemplate{footline}[text line]{}
\setbeamercolor{structure}{fg=purple!50!blue, bg=purple!50!blue}

\setbeamersize{text margin left=15pt,text margin right=15pt}

\setbeamercovered{transparent}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\makeatother

\begin{document}

\title{A Discrepancy-based Framework to Compare Robustness between Multi-Attribute Evaluations}

\author{J.~Raimbault$^{1,2}$\\
\texttt{juste.raimbault@parisgeo.cnrs.fr}
}

\institute{$^{1}$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
$^{2}$UMR-T IFSTTAR 9403 LVMT\\
}

\date{CSDM 2016 - Paris\\\smallskip
14th December 2016
}

\frame{\maketitle}

%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%

\sframe{Complex systems}{

% introduction with pictures/anecdote

}

\sframe{Multi-attribute Evaluation in Complex Socio-technical Systems}{

\justify

}

\sframe{Robustness of Evaluations}{

\justify

}

\sframe{Towards a Framework}{

\justify

\textbf{Research Objective : } \textit{}

% put rationales here

\bigskip

$\rightarrow$ 
\medskip

$\rightarrow$

}

%%%%%%%%%%%%%%%%%
\section{Framework Description}
%%%%%%%%%%%%%%%%%

\sframe{Intuitive Description}{

}

\sframe{Formal Description (I)}{

}

\sframe{Formal Description (II)}{

}

%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%

\sframe{Implementation on Synthetic Data}{

% summary, data description, main result (summary of table)

}

\sframe{Metropolitan Segregation}{

% indicators, data

}

\sframe{Metropolitan Segregation}{

% segreg maps

}

\sframe{Metropolitan Segregation}{

% missing data

}

%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%

\sframe{Applicability}{

\justify

}

\sframe{Possible Developments}{

\justify

}

\sframe{Conclusion}{

$\rightarrow$

\medskip

\bigskip
\bigskip
\bigskip

\footnotesize{ - All code available at \texttt{https://github.com/JusteRaimbault/RobustnessDiscrepancy}

\medskip

 - Paper preprint available at \texttt{http://arxiv.org/abs/}
}

}

%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{/Users/Juste/Documents/ComplexSystems/CityNetwork/Biblio/Bibtex/CityNetwork,biblio}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2016-12-11 Docs/Communication/discrepancy.tex
slides

@@
 
+%%%
+% Abstract
+
+%Multi-objective evaluation is a necessary aspect when managing complex systems, as the intrinsic complexity of a system is generally closely linked to the potential number of optimization objectives.
+However, an evaluation makes no sense without its robustness being given (in the sense of its reliability).
+Statistical robustness computation methods are highly dependent of underlying statistical models.
+We propose a formulation of a model-independent framework in the case of integrated aggregated indicators (multi-attribute evaluation), that allows to define a relative measure of robustness taking into account data structure and indicator values.
+We implement and apply it to a synthetic case of urban systems based on Paris districts geography, and to real data for evaluation of income segregation for Greater Paris metropolitan area.
+First numerical results show the potentialities of this new method.
+Furthermore, its relative independence to system type and model may position it as an alternative to classical statistical robustness methods.
+%\keywords{Multi-attribute Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
+
 %%%%%%%%%%%%%%%%%
@@
 
+\textit{Systematic multi-objective nature of problems in design of Complex Industrial Systems~\cite{marler2004survey} and in the study of Complex Natural Systems~\cite{newman2011complex}}
+
+\medskip
+
+$\rightarrow$
+
+\medskip
+
+$\rightarrow$ Territorial systems as typical examples : e.g.
+sustainable urban design~\cite{souami2012ecoquartiers}, multi-criteria decision-making for transportation infrastructures~\cite{bavoux2005geographie}
+
 }
@@
 
+.
+
 }
 
-\sframe{Towards a Framework}{
+\sframe{Towards a Generic Robustness Framework}{
 
@@
 
-\textbf{Research Objective : } \textit{}
+\textbf{Research Objective : } \textit{Investigate a generic data-driven approach to Robustness in Multi-attribute evaluations of Complex Socio-Technical Systems}
 
@@
 
-\sframe{Intuitive Description}{
+\sframe{Theoretical Framing}{
+
+% here theory is in the sense of thematic theory
+
+}
+
+\sframe{Assumptions}{
+\justify
+
+% assumptions done
+
+\textbf{Objectives as Spatial Integrals}
+
+\medskip
+
+\textbf{Linearly Aggregated Objectives}
 
@@
 
+Territorial Systems $S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}$ with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$
+
+\[
+(\mathcal{X},\mathcal{Y}) \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right) = \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
+\]
+
+Objectives : $H_{c}$ space of real-valued functions on $(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})$, such that for all $h\in H_{c}$ :
+\begin{enumerate}
+\item $h$ is ``enough'' regular (tempered distributions e.g.)
+\item $q_c=\int_{(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})}h$ is a function describing the ``urban fact'' (the indicator in itself)
+\end{enumerate}
+
 }
@@
 \sframe{Formal Description (II)}{
+
+Integral approximation theorem gives upper bound on error, linked to data discrepancy~\cite{niederreiter1972discrepancy}\cite{varet2010developpement}
+
+\[
+\left\Vert \int h_{c}-\frac{1}{n_{i,c}}\sum_{l}h_{c}(\vec{X}_{i,c,l})\right\Vert \leq K\cdot\left|\left|\left|h_{c}\right|\right|\right|\cdot D_{i,c}
+\]
+
+which propagates to the linear aggregation
+
+\[
+\left\Vert \int\sum w_{i,c}h_{c}-\frac{1}{n_{i,c}}\sum_{l}w_{i,c}h_{c}(\vec{X}_{i,c,l})\right\Vert \leq K\sum_{c}\left|w_{i,c}\right|\left|\left|\left|h_{c}\right|\right|\right|\cdot D_{i,c}
+\]
+
+}
+
+\sframe{Formal Description (III)}{
+
+A relative \textit{Robustness Ratio} can thus be defined between two evaluations :
+
+\begin{equation}
+R_{i,i'}=\frac{\sum_{c}w_{i,c}\cdot D_{i,c}}{\sum_{c}w_{i',c}\cdot D_{i',c}}
+\end{equation}
 
@@
 
+Metropolitan Segregation on Ile-de-France, Insee income data (2011)
+
+\medskip
+
+\textbf{Indicators : }
+
+\begin{itemize}
+\item Spatial autocorrelation Moran index, defined as weighted normalized covariance of median income by $\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}$
+\item Dissimilarity index (close to Moran but integrating local dissimilarities rather than correlations), given by $d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|$\\ with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
+\item Complementary of the entropy of income distribution that is a way to capture global inequalities $\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)$
+\end{itemize}
+
 }
@@
 
+\textit{Example of Segregation maps}
+
+\medskip
+
+\includegraphics[width=\textwidth]{figures/grandParis_income_moran}
+
 }
@@
 % missing data
+\textit{Framework Application : sensitivity to missing data}
+
+\medskip
+
+\includegraphics[width=0.45\textwidth]{figures/alldeps_rob_renormindics}
+
+\includegraphics[width=0.45\textwidth]{figures/alldeps_robsd_renormindics}
+
+% interpretation : different regimes.
 
@@
 
+$\rightarrow$ Application to decision-making procedures : adding robustness as a dimension ?
+
+\medskip
+
+$\rightarrow$ Availability of raw data
+
+\medskip
+
+$\rightarrow$ Assumptions validity ranges : some indicators may difficultly be viewed as spatial integrals (as some accessibility measures~\cite{kwan1998space}
+
 }
 
-\sframe{Possible Developments}{
+\sframe{Further Developments}{
 
 \justify
+
+$\rightarrow$ Application to existing open frameworks (e.g. \cite{tivadar2014oasis})
+
+\medskip
+
+$\rightarrow$ More general formulation, first to non-linear aggregation (e.g. for Lipschitzian functions~\cite{dragomir1999ostrowski})
 

### 2016-12-11 Docs/Communication/discrepancy.tex
slides_v1

@@
 
-\sframe{Complex systems}{
+\subsection{Context}
+
+\sframe{Multi-objective Complex Systems}{
 
@@
 
+\textit{From morphogenetic factors to optimal design, fundamental multi-objective nature of optimization processes}
+
+\bigskip
+
+\includegraphics[width=0.35\textwidth]{figures/lattice}\hspace{0.5cm}
+\includegraphics[width=0.55\textwidth]{figures/exampleOptimalCorridor}
+
+\smallskip
+
+{\footnotesize Source : Left \cite{raimbault2014hybrid} ; Right \cite{raimbault2015labex}}
+
 }
 
-\sframe{Multi-attribute Evaluation in Complex Socio-technical Systems}{
+\sframe{Multi-objective Evaluations of Socio-technical Systems}{
 
@@
 
-\medskip
+\bigskip
+\bigskip
 
-$\rightarrow$
+$\rightarrow$ Design and understanding (i.e. evaluation) of Socio-technical systems at the intersection as hybrid systems~\cite{picon2013smart}\cite{haken2003face}
 
-\medskip
+\bigskip
 
@@
 
-.
+\textit{Reliability of an evaluation is crucial; various approaches to robustness depending on method}
+
+\bigskip
+\bigskip
+
+$\rightarrow$ Naturally included in the construction and estimation of statistical models~\cite{launer2014robustness} (e.g. p-value, beta power, AIC)
+
+%\bigskip
+
+%$\rightarrow$ Bayesian statistics also include robustness of sequential estimation
+
+\bigskip
+
+$\rightarrow$ In multi-objective optimization, diverse methods: sensitivity of Pareto front to perturbations~\cite{deb2006introducing};  continuity of solutions
+
+\cite{1688537}
+
+\bigskip
+
+$\rightarrow$ \cite{dobbie2013robustness} studies robustness for multi-attribute evaluations as biases depending on weighting techniques
 
 }
+
+\subsection{Research Question}
 
@@
 
+% -> why multi -attribute
+% -> data-driven
+% -> model-independant
+
+\bigskip
 \bigskip
 
-$\rightarrow$ 
+$\rightarrow$ Model-independence and method-independence; framework based only on data structure (and thus quality) and indicator values
 \medskip
 
-$\rightarrow$
+$\rightarrow$ Particular case of multi-attribute evaluations, where dimensions are aggregated, in order to obtain a simple measure of robustness
 
@@
 
-\sframe{Theoretical Framing}{
+\subsection{Theoretical Framing}
+
+\sframe{Intuitive and Theoretical Framing}{
 
@@
 
+\justify
+
+\begin{enumerate}
+\item {\justify Systems are seen from the perspective of raw data available: data-driven approach}
+
+\smallskip
+
+\item  A choice of indicators captures the realization of an ``urban fact''~\cite{mangin1999projet}, in the sense of stylized process with different spatial realizations
+
+\smallskip
+
+\item Given many systems and indicators, a common space can be build to compare them
+
+\smallskip
+
+\item Discrepancy of data~\cite{dick2010digital} in that space captures various aspects linked to robustness: system scale and range, precision of data, missing data
+
+\smallskip
+
+\item Robustness must also capture an error done on indicator computation
+
+\end{enumerate}
+
 }
+
+\subsection{Assumptions}
 
@@
 
-\textbf{Objectives as Spatial Integrals}
+\textbf{Objectives as Kernel Integrals}
 
@@
 
+$\rightarrow$ Kernel functions exist for each objective, computed as their integrals
+
+\medskip
+
+$\rightarrow$ Reasonable for most systems, as e.g. for territorial systems  it is analog to smoothing of Geographically Weighted Regression~\cite{brunsdon1998geographically}
+
+\bigskip
+
 \textbf{Linearly Aggregated Objectives}
+
+\medskip
+
+$\rightarrow$ Aggregated objective as linear combination of attributes $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$
+
+\medskip
+
+$\rightarrow$ Choice of weights at the core of decision-making process; not our scope here, we take simply relative indicator importance $w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}$
 
@@
 
+\subsection{Formal Description}
+
 \sframe{Formal Description (I)}{
 
-Territorial Systems $S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}$ with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$
+\textbf{Territorial Systems : } Data $S_{i}=\mathbf{X}_{i}\in\mathcal{X}_{i}$ with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$
+
+\medskip
+
+\textbf{System space}
 
 \[
-(\mathcal{X},\mathcal{Y}) \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right) = \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
+\mathcal{X} \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right) = \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)
 \]
 
-Objectives : $H_{c}$ space of real-valued functions on $(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})$, such that for all $h\in H_{c}$ :
+\medskip
+
+\textbf{Objectives :} $H_{c}$ space of real-valued functions on $\tilde{\mathcal{X}}_{c}$, such that :
 \begin{enumerate}
-\item $h$ is ``enough'' regular (tempered distributions e.g.)
-\item $q_c=\int_{(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})}h$ is a function describing the ``urban fact'' (the indicator in itself)
+\item $h_c\in H_{c}$ are ``enough'' regular (tempered distributions e.g.)
+\item $q_c=\int_{\tilde{\mathcal{X}}_{c}}h$ is a function describing the ``urban fact'' (the indicator in itself)
+\item Normalized kernels $h_c (\vec{x}) \in [0,1]$ % TYPO PAPER HERE ? not clear beween h_c and q_c ; both can be and are in applications
 \end{enumerate}
@@
 
-}
+\bigskip
 
-\sframe{Formal Description (III)}{
-
-A relative \textit{Robustness Ratio} can thus be defined between two evaluations :
+$\rightarrow$ A relative \textit{Robustness Ratio} can thus be defined between two evaluations :
 
@@
 %%%%%%%%%%%%%%%%%
-\section{Results}
+\section{Application}
 %%%%%%%%%%%%%%%%%
+
+\subsection{Synthetic Data}
 
@@
 
+\textit{Using OpenStreetMap roads and buildings data, construction of synthetic indicators on Paris districts (car daily use, car flows in streets, relative length of pedestrian streets), computed in Monte-carlo simulations}
+
+\medskip
+
+$\rightarrow$ Minimal ratio (relative to 1st Arr.) in 15th Arr. ($0.92 \pm 0.03$), intuitively expected
+
+\bigskip
+
+\centering
+
+\includegraphics[width=0.5\textwidth]{figures/ex_data}
+
+\smallskip
+
+\small \textit{Example of raw data}
+
 }
+
+\subsection{Metropolitan Segregation}
 
@@
 
-Metropolitan Segregation on Ile-de-France, Insee income data (2011)
+\textit{Application to metropolitan Segregation on Ile-de-France, Insee income data (2011)}
 
-\medskip
+\bigskip
+\bigskip
 
@@
 \begin{itemize}
-\item Spatial autocorrelation Moran index, defined as weighted normalized covariance of median income by $\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}$
-\item Dissimilarity index (close to Moran but integrating local dissimilarities rather than correlations), given by $d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|$\\ with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
-\item Complementary of the entropy of income distribution that is a way to capture global inequalities $\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)$
+\item Spatial autocorrelation Moran index
+%defined as weighted normalized covariance of median income by $\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}$
+\item Dissimilarity index $d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|$
+%with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
+\item Complementary of distribution entropy $\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)$
 \end{itemize}
@@
 
-\sframe{Metropolitan Segregation}{
+\sframe{Metropolitan Segregation: Indicators}{
 
@@
 
-\textit{Example of Segregation maps}
+\textit{Example of Segregation maps (linking median income with local Moran index)}
 
@@
 
-\sframe{Metropolitan Segregation}{
+\sframe{Metropolitan Segregation: Results}{
 
 % missing data
-\textit{Framework Application : sensitivity to missing data}
+\textit{Framework Application : robustness comparison between administrative areas and sensitivity to missing data}
+
+\bigskip
+
+\centering
+
+\includegraphics[width=0.55\textwidth]{figures/alldeps_rob_renormindics}
+\includegraphics[width=0.55\textwidth]{figures/alldeps_robsd_renormindics}
+
+% interpretation : different regimes.
+
+\small \textit{(Left) Robustness ratio compared to full region for each area, as a function of missing data proportion; (Right) Standard deviations}
+
+}
+
+\subsection{Discussion}
+
+\sframe{Discussion}{
+
+\justify
+
+\vspace{-0.5cm}
+
+\textbf{Applicability}
 
@@
 
-\includegraphics[width=0.45\textwidth]{figures/alldeps_rob_renormindics}
+$\rightarrow$ Application to decision-making procedures : adding robustness as a dimension
 
-\includegraphics[width=0.45\textwidth]{figures/alldeps_robsd_renormindics}
+\smallskip
 
-% interpretation : different regimes.
+$\rightarrow$ Assumptions validity ranges : some indicators may difficultly be viewed as spatial integrals (as some accessibility measures~\cite{kwan1998space})
 
-}
+\smallskip
 
-%%%%%%%%%%%%%%%%%
-\section{Discussion}
-%%%%%%%%%%%%%%%%%
+$\rightarrow$ Availability of raw data
 
-\sframe{Applicability}{
+\bigskip
 
-\justify
-
-$\rightarrow$ Application to decision-making procedures : adding robustness as a dimension ?
+\textbf{Further Developments}
 
@@
 
-$\rightarrow$ Availability of raw data
-
-\medskip
-
-$\rightarrow$ Assumptions validity ranges : some indicators may difficultly be viewed as spatial integrals (as some accessibility measures~\cite{kwan1998space}
-
-}
-
-\sframe{Further Developments}{
-
-\justify
-
 $\rightarrow$ Application to existing open frameworks (e.g. \cite{tivadar2014oasis})
 
-\medskip
+\smallskip
 
 $\rightarrow$ More general formulation, first to non-linear aggregation (e.g. for Lipschitzian functions~\cite{dragomir1999ostrowski})
+
+\smallskip
+
+$\rightarrow$ Multi-dimensional measure of robustness using discrepancy along different dimensions
 
@@
 \sframe{Conclusion}{
+\justify
 
-$\rightarrow$
+$\rightarrow$ An original data-driven approach to robustness in multi-attribute evaluations of complex socio-technical systems; independent of model and method
 
@@
 
-\bigskip
-\bigskip
-\bigskip
-
-\footnotesize{ - All code available at \texttt{https://github.com/JusteRaimbault/RobustnessDiscrepancy}
+$\rightarrow$ Future insights from application to diverse system types and fields should refine the framework
 
@@
 
- - Paper preprint available at \texttt{http://arxiv.org/abs/}
+$\rightarrow$ Importance of interdisciplinarity (linking here statistics with computational modeling), crucial to study Complex Systems
+
+\bigskip
+
+\footnotesize{
+
+- Special thanks to J. Keutchayan (Ecole Polytechnique de Montr{\'e}al) for suggesting the original idea of using discrepancy
+
+\medskip
+
+- All code and data available at \texttt{https://github.com/JusteRaimbault/RobustnessDiscrepancy}
+
+\medskip
+
+ - Paper preprint available \cite{raimbault2016discrepancy} at \texttt{http://arxiv.org/abs/1608.00840}
 }
@@
 }
+
+\sframe{Reserve Slides}{
+
+\centering
+
+\Large \textbf{Reserve Slides}
+
+}
+
+\sframe{Kernel Examples}{
+
+Typical concrete example of kernels can be :
+
+\begin{itemize}
+\item A mean of rows of $\mathbf{X}_{i,c}$ is computed with $h(x)=x\cdot f_{i,c}(x)$ where $f_{i,c}$ is the density of the distribution of the assumed underlying variable.
+\item A rate of elements respecting a given condition $C$, $h(x)=f_{i,c}(x)\chi_{C(x)}$ 
+\item For already aggregated variables $\mathbf{Y}$, a Dirac distribution allows to express them also as a kernel integral. 
+\end{itemize}
+
+}
+
+\sframe{Implementation}{
+
+\begin{itemize}
+\item Preprocessing of geographical data is made through QGIS
+
+\cite{qgis2011quantum}
+\item Core implementation of the framework is done in R \cite{team2000r} for the flexibility of data management and statistical computations 
+\item package \texttt{DiceDesign}\cite{franco20092} for computation of discrepancies
+\end{itemize}
+
+}
+
+\sframe{Synthetic Indicators (I)}{
+\begin{itemize}
+\item Complementary of the average daily distance to work with car per individual, approximated by, with $n_{cars}(b)$ number of cars in the building (randomly generated by associated of cars to a number of building proportional to motorization rate $\alpha_m ~ 0.4$ in Paris), $d_w$ distance to work of individuals (generated from the building to a uniformly generated random point in spatial extent of the dataset), and $d_{max}$ the diameter of Paris area, $\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}$
+
+\end{itemize}
+
+}
+
+\sframe{Synthetic Indicators (II)}{
+
+\begin{itemize}
+\item Complementary of average car flows within the streets in the district, approximated by, with $\varphi(s)$ relative flow in street segment $s$, generated through the minimum of 1 and a log-normal distribution adjusted to have $95\%$ of mass smaller than 1 what mimics the hierarchical distribution of street use (corresponding to betweenness centrality), and $l(s)$ segment length, $\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}$
+\item Relative length of pedestrian streets $\bar{p}$, computed through a randomly uniformly generated dummy variable adjusted to have a fixed global proportion of segments that are pedestrian.
+\end{itemize}
+}
+
+\begin{frame}
+\frametitle{Synthetic Numerical Results}
+
+\includegraphics[width=\textwidth,height=0.9\textheight]{figures/syntheticnum.png}
+
+\end{frame}
 

### 2016-12-19 Docs/translation/paper_FR.tex
trad

--- 2016-08-08 Docs/paper/paper.tex
@@
 % first the title is needed
-\title{A Discrepancy-based Framework to Compare Robustness between Multi-Attribute Evaluations}
+\title{Un Cadre Basé sur la Discrépance pour une Comparaison de la Robustesse entre Evaluations Multi-attributs}
 
 % a short form should be given in case it is too long for the running head
-\titlerunning{Discrepancy-based Framework}
+\titlerunning{Un Cadre Basé sur la Discrépance}
 
@@
 %
-\author{\noun{Juste Raimbault}$^{1,2}$} %and \noun{Julien Keutchayan}$^{3}$}
+\author{\noun{Juste Raimbault}$^{1,2}$} 
 %
-\authorrunning{Discrepancy-based Framework}
+\authorrunning{Un Cadre Basé sur la Discrépance}
 % (feature abused for this document to repeat the title also on left hand pages)
@@
 $^{2}$ UMR-T IFSTTAR 9403 LVMT, Champs-sur-Marne, France\\
-%$^{3}$ D{\'e}partement de math{\'e}matiques et de g{\'e}nie industriel,\\
-%Ecole Polytechnique de Montr{\'e}al, Montr{\'e}al, Canada\\\medskip
 \mailjr\\
-%\mailjk
 }
@@
 \begin{abstract}
-Multi-objective evaluation is a necessary aspect when managing complex systems, as the intrinsic complexity of a system is generally closely linked to the potential number of optimization objectives.
-%In the case of spatial socio-technical systems (territorial and urban systems), the recent example of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where such an evaluation is central to the development of sustainable innovative concepts.
-However, an evaluation makes no sense without its robustness being given (in the sense of its reliability).
-Statistical robustness computation methods are highly dependent of underlying statistical models.
-We propose a formulation of a model-independent framework in the case of integrated aggregated indicators (multi-attribute evaluation), that allows to define a relative measure of robustness taking into account data structure and indicator values.
-We implement and apply it to a synthetic case of urban systems based on Paris districts geography, and to real data for evaluation of income segregation for Greater Paris metropolitan area.
-First numerical results show the potentialities of this new method.
-Furthermore, its relative independence to system type and model may position it as an alternative to classical statistical robustness methods.
-\keywords{Multi-attribute Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
+Les évaluations multi-objectifs sont un aspect essentiel de la gestion de systèmes complexes, puisque la complexité intrinsèque d'un système est généralement étroitement liée au nombre d'objectifs d'optimisation potentiels.
+Cependant, une évaluation ne fait pas sens si sa robustesse, au sens de sa fiabilité, n'est pas donnée.
+Les méthodes statistiques usuelles fournissant une mesure de robustesse sont très dépendantes des modèles sous-jacents.
+Nous proposons une formulation d'un cadre indépendant du modèle, dans le cas d'indicateurs intégrés et agrégés (évaluation multi-attributs), qui permet de définir une mesure de robustesse relative prenant en compte la structure des données et les valeurs des indicateurs.
+La méthode est testée sur données urbaines synthétiques associées aux arrondissements de Paris, et à des données réelles de revenus pour l'évaluation de la ségrégation urbaine dans la région métropolitaine du Grand Paris.
+Les premiers résultats numériques montrent les potentialités de cette nouvelle méthode.
+De plus, sa relative indépendance au type de système et au modèle pourrait la positionner comme une alternative aux méthodes statistiques classiques d'évaluation de la robustesse.
+\keywords{Evaluation Multi-attributs, Robustesse Non-modèle-dépendante, Systèmes Urbains, Discrépance}
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%
-\subsection{General Context}
+\subsection{Contexte Général}
 
-Multi-objective problems are organically linked to the complexity of underlying systems.
-Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration often leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of non engineered physical, biological or social systems that exhibit emergence and self-organization properties, where objectives can e.g.
-be the result of heterogeneous interacting agents (see~\cite{newman2011complex} for a large survey of systems concerned by this approach), multi-objective optimization can be explicitly introduced to study or design the system but is often already implicitly ruling the internal mechanisms of the system.
-The case of socio-technical Complex Systems is particularly interesting as, following~\cite{haken2003face}, they can be seen as hybrid systems embedding social agents into ``technical artifacts'' (sometimes to an unexpected degree creating what \noun{Picon} describes as \emph{cyborgs}~\cite{picon2013smart}), and thus cumulate propensity to be at the origin of multi-objective issues\footnote{We design by \emph{Multi-Objective Evaluation} all practices including the computation of multiple indicators of a system (it can be multi-objective optimization for system design, multi-objective evaluation of an existing system, multi-attribute evaluation ; our particular framework corresponds to the last case).}.
-The new notion of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where sustainability implies contradictory objectives.
-The example of transportation systems, which conception shifted during the second half of the 20th century from cost-benefit analysis to multi-criteria decision-making, is also typical of such systems~\cite{bavoux2005geographie}.
-Geographical system are now well studied from such a point of view in particular thanks to the integration of multi-objective frameworks within Geographical Information Systems~\cite{carver1991integrating}.
-As for the micro-case of eco-districts, meso and macro urban planning and design may be made sustainable through indicators evaluation~\cite{jegou2012evaluation}.
+Les problèmes multi-objectifs sont organiquement liés à la complexité des systèmes sous-jacents.
+En effet, que ce soit dans le champ des \emph{Systèmes Complexes Industriels}, dans le sens de systèmes conçus par ingénierie, où la construction de Systèmes de Systèmes (SoS) par couplage et intégration induit souvent des objectifs contradictoires~\cite{marler2004survey}, ou dans le champ des \emph{Systèmes Complexes Naturels}, au sens de systèmes non désignés, physiques, biologiques ou sociaux, qui présentent des propriétés d'émergence et d'auto-organisation, pour lesquels les objectifs peuvent e.g.
+être le résultat de l'interaction d'agents hétérogènes (voir~\cite{newman2011complex} pour une revue étendue des types de systèmes concernés par cette approche), l'optimisation multi-objectifs peut être explicitement introduite pour étudier ou désigner le système, mais régit généralement déjà implicitement les mécanismes internes du système.
+Le cas des Systèmes Complexes Sociaux-techniques est particulièrement intéressant puisque selon Haken~\cite{haken2003face}, ils peuvent être vus comme des systèmes hybrides embarquant des agents sociaux dans des ``artefacts techniques'' (parfois jusqu'à un niveau inattendu, créant ce que \noun{Picon} décrit comme \emph{cyborgs}~\cite{picon2013smart}), et cumulent ainsi la potentialité d'être à l'origine de problèmes multi-objectifs\footnote{Nous désignons ici par \emph{Evaluation Multi-objectifs} toutes les pratiques incluant le calcul de multiples indicateurs d'un système (il peut s'agir d'optimisation multi-objectif pour un design de système, une évaluation multi-objectif d'un système existant, une évaluation multi-attributs ; notre cadre particulier correspondra au dernier cas).}.
+La notion récente d'\emph{éco-quartier}~\cite{souami2012ecoquartiers} est un exemple typique pour lequel la durabilité implique des objectifs contradictoires.
+L'exemple des systèmes de transport, dont la conception a glissé durant la seconde moitié du 20ème siècle d'analyses coût-bénéfices à la price de décision multi-critères, est également typique de tels systèmes~\cite{bavoux2005geographie}.
+Les systèmes géographiques sont à présent bien étudiés d'un tel point de vue, en particulier grâce à l'intégration des cadres multi-objectifs au sein des Systèmes d'Information Géographiques~\cite{carver1991integrating}.
+Comme dans le cas microscopique des éco-quartiers, la planification et le design urbains mésoscopiques et macroscopiques peuvent être rendus durables grâce aux évaluations par indicateurs~\cite{jegou2012evaluation}.
 
-A crucial aspect of an evaluation is a certain notion of its reliability, that we call here \emph{robustness}.
-% Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework.
-Statistics naturally include this notion since the construction and estimation of statistical models give diverse indicators of the consistence of results~\cite{launer2014robustness}.
-The first example that comes to mind is the application of the law of large numbers to obtain the \emph{p-value} of a model fit, that can be interpreted as a confidence measure of estimates.
-Besides, confidence intervals and \emph{beta-power} are other important indicators of statistical robustness.
-Bayesian inference provide also measures of robustness when distribution of parameters are sequentially estimated.
-Concerning multi-objective optimization, in particular through heuristic algorithms (for example genetic algorithms, or operational research solvers), the notion of robustness of a solution concerns more the stability of the solution on the phase space of the corresponding dynamical system.
-Recent progresses have been done towards unified formulation of robustness for a multi-objective optimization problem, such as~\cite{deb2006introducing} where robust Pareto-front as defined as solutions that are insensitive to small perturbations.
-In~\cite{1688537}, the notion of degree of robustness is introduced, formalized as a sort of continuity of other solutions in successive neighborhood of a solution.
+% TODO Pour une théorie des systèmes sociaux-techniques : un des axes majeurs de développement ?
 
-However, there still lack generic methods to estimate robustness of an evaluation that would be model-independent, i.e.
-that would be extracted from data structure and indicators but that would not depend on the method used.
-Some advantages could be for example an \emph{a priori} estimation of potential robustness of an evaluation and thus to decide if the evaluation is worth doing.
-We propose here a framework answering this issue in the particular case of Multi-attribute evaluations, i.e.
-when the problem is made unidimensional by objectives aggregation.
-It is data-driven and not model-driven in the sense that robustness estimation does not depend on how indicators are computed, as soon as they respect some assumptions that will be detailed in the following.
+Un aspect crucial de l'évaluation est une certaine notion de sa fiabilité, que nous nommerons ici \emph{robustesse}.
+Les méthodes statistiques incluent naturellement cette notion puisque la construction et l'estimation de modèles statistiques donne divers indicateurs de la consistence des résultats~\cite{launer2014robustness}.
+Le premier exemple venant à l'esprit est l'application de la loi des grands nombres pour obtenir la \emph{p-valeur} d'une estimation de modèle, qui peut être interprété comme une mesure de confiance en les valeurs estimées.
+D'autre part, les intervalles de confiance et le  \emph{beta-power} sont d'autres indicateurs importants de robustesse statistique.
+L'inférence bayésienne fournit également des mesures de robustesse quand la distribution des paramètres est estimée de manière séquentielle.
+Concernant les optimisations multi-objectifs, en particulier par des algorithmes heuristiques (comme par exemple les algorithmes génétiques, ou les solveurs de recherche opérationelle), la notion de robustesse d'une solution consiste plus en la stabilité de la solution dans l'espace des phases du système dynamique correspondant.
+Des progrès récents ont été faits vers une formulation unifiée de la robustesse pour les problèmes d'optimisation multi-objectifs, comme dans~\cite{deb2006introducing} où les fronts de Pareto robustes sont définis comme des solutions insensibles aux petites perturbations.
+Dans~\cite{1688537}, la notion de degré de robustesse est introduite, formalisée comme une sorte de continuité des autres solutions dans des voisinages successifs d'une solution.
+
+Cependant, il n'existe pas de méthode générique qui permettrait une évaluation de la robustesse de façon indépendante au modèle, i.e.
+qui serait extraite de la structure des données et des indicateurs mais ne dépendrait pas de la méthode utilisée.
+Un avantage serait par exemple une estimation \emph{a priori} de la robustesse potentielle d'une évaluation et de décider ainsi si elle vaut la peine d'être faite.
+Nous proposons un cadre répondant à cette contrainte dans le cas particulier des évaluations multi-attributs, i.e.
+quand le problème est rendu unidimensionnel par agrégation des objectifs.
+Il est basé sur les données et non sur les modèles, au sens ou l'estimation de la robustesse ne dépendra pas de la manière dont les indicateurs sont calculés, tant qu'ils respectent certaines hypothèses détaillées par la suite.
 
 %%%%%%%%%%%%%%%%
-\subsection{Proposed Approach}
+\subsection{Approche Proposée}
 
-\paragraph{Objectives as Spatial Integrals}
+\paragraph{Objectifs comme Intégrales Spatiales}
 
-%Our approach needs to restrict the scope of application.
-We assume that objectives can be expressed as spatial integrals, so it should apply to any territorial system and our application cases are urban systems.
-It is not that restrictive in terms of possible indicators if one uses suitable variables and integrated kernels : in a way analog to the method of geographically weighted regression~\cite{brunsdon1998geographically}, any spatial variable can be integrated against regular kernels of variable size and the result will be a spatial aggregation which sense depends on kernel size.
-The example we use in the following such as conditional means or sums suit well the assumption.
-Even an already spatially aggregated indicator can be interpreted as a spatial indicator by using a Dirac distribution on the centroid of the corresponding area.
+Nous supposons que les objectifs peuvent être exprimés comme intégrales spatiales, ce qui devrait s'appliquer à tout système territorial, et nos cas d'application sont des systèmes urbains.
+Ce n'est pas si restrictif en terme d'indicateurs possibles si l'on utilise les bonnes variables et noyaux intégrés : de façon analogue à la méthode de Regression Géographique Pondérée~\cite{brunsdon1998geographically}, toute variable spatiale peut être intégrée contre des noyaux réguliers de taille variable et le résultats sera une agrégation spatiale dont la signification dépendra de l'étendue du noyau.
+Les exemples utilisés par la suite comme des moyennes conditionnelles ou des sommes vérifient parfaitement cette hypothèse.
+Même un indicateur déjà agrégé dans l'espace peut être interprété comme une intégrale spatiale en utilisant une distribution de Dirac au centroïde de la zone correspondante.
 
-\paragraph{Linearly Aggregated Objectives}
+\paragraph{Objectifs Agrégés Linéairement}
 
@@
 It is analog to aggregated utility techniques in economics and is used in many fields.
-%many domains apply such a method that can be called \emph{Multi-Criteria Decision Analysis}~\cite{wang2009review}.
 The subtlety lies in the choice of weights, i.e.
@@
 
-The rest of the paper is organized as follows : section 2 describes intuitively and mathematically the proposed framework ; section 3 then details implementation, data collection for case studies and numerical results for an artificial intra-urban case and a metropolitan real case ; section 4 finally discuss limitations and potentialities of the method.
+Le reste de cette monographie est organisé de la façon suivante : la section 2 décrit intuitivement puis mathématiquement le cadre proposé ; la section 3 détaille ensuite l'implémentation, la collecte des données pour les cas d'étude et les résultats numériques pour une évaluation intra-urbaine synthétique et un cas réel métropolitain ; la section 4 discute finalement les limitations et les potentialités de la méthode.
 

### 2016-12-22 Docs/translation/paper_FR.tex
more_tr

@@
 \newcommand{\noun}[1]{\textsc{#1}}
+
+% encoding 
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 
@@
 
-A second assumption we make is that the multi-objective evaluation is done through linear aggregation of objectives, i.e.
-that we are tackling a multi-attribute optimization problem.
-If $(q_i(\vec{x}))_i$ are values of objectives functions, then weights $(w_i)_i$ are defined in order to build the aggregated decision-making function $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, which value determines then the performance of the solution.
-It is analog to aggregated utility techniques in economics and is used in many fields.
-The subtlety lies in the choice of weights, i.e.
-the shape of the projection function, and various approaches have been developed to find weights depending on the nature of the problem.
-Recent work~\cite{dobbie2013robustness} proposed to compare robustness of different aggregation techniques through sensitivity analysis, performed by Monte-Carlo simulations on synthetic data.
-Distribution of biases where obtained for various techniques and some showed to perform significantly better than others.
-Robustness assessment still depended on models used in that work.
+Une seconde hypothèse que nous faisons est que l'évaluation multi-objectifs est effectuée par agrégation linéaire des objectif, c'est à dire qu'on se place dans le cadre d'un problème d'optimisation multi-attributs.
+Si $(q_i(\vec{x}))_i$ sont les valeurs des fonctions objectifs, on définit alors des poids $(w_i)_i$ afin de construire la fonction de prise de décision $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, dont la valeur détermine ensuite la performance d'une solution.
+Cette approche est analogue aux utilités agrégées en économie et est utilisée dans de nombreux domaines.
+La subtilité réside dans le choix des poids, i.e.
+de la forme de la fonction de projection, et différentes solutions ont été dévelopées pour obtenir des poids selon la nature du problème.
+Récemment, \cite{dobbie2013robustness} a proposé de comparer la robustesse des différentes techniques d'agrégation par une analyse de sensibilité, effectuée par simulations de Monte-Carlo pour produire des données synthétiques, ce qui permet d'obtenir la distribution des biais pour les différentes techniques, certaines étant significativement plus performantes que d'autres.
+Toutefois, la quantification de la robustesse dépend toujours des modèles utilisés dans ce travail.
 
@@
 %%%%%%%%%%%%%%%%
-\section{Framework Description}
+\section{Description du Cadre}
 
 %%%%%%%%%%%%%%%%
-\subsection{Intuitive Description}
+\subsection{Description Intuitive}
 
-We describe now the abstract framework allowing theoretically to compare robustnesses of evaluations of two different urban systems.
-Our framework is a generalization of an empirical method proposed in~\cite{ecodistrictReport} besides a more general benchmarking study on indicator sense and relevance in a sustainability context.
-Intuitively, it relies on empirical base resulting from the following axioms :
+Nous décrivons à présent le cadre proposé pour permettre théoriquement de comparer la robustesse d'évaluation de deux systèmes urbains différents.
+Ce cadre est une généralisation d'une méthode empirique proposée dans~\cite{ecodistrictReport} pour accompagner une étude dans un autre contexte effectuant une comparaison du sens et de la pertinence des indicateurs dans un contexte de durabilité.
+Intuitivement, la base empirique se base sur les principes suivants :
+
 \begin{itemize}
-\item Urban systems can be seen from the information available, i.e.
-raw data describing the system.
-As a data-driven approach, this raw data is the basis of our framework and robustness will be determined by its structure.
-\item From data are computed indicators (objective functions).
-We assume that a choice of indicators is an intention to translate particular aspects of the system, i.e.
-to capture a realization of an ``urban fact'' (\emph{fait urbain}) in the sense of \noun{Mangin}~\cite{mangin1999projet} - a sort of stylized fact in terms of processes and mechanisms, having various realizations on spatially distinct systems, depending on each precise context.
-\item Given many systems and associated indicators, a common space can be built to compare them.
-In that space, data represents more or less well real systems, depending e.g.
-on initial scale, precision of data, missing data.
-We precisely propose to capture that through the notion of point cloud discrepancy, which is a mathematical tool coming from sampling theory expressing how a dataset is distributed in the space it is embedded in~\cite{dick2010digital}.
-%Coupling discrepancies with appropriate weights depending on indicator importance allows to introduce a relative ratio of robustness between two evaluations.
+\item Les systèmes urbains peuvent être vus selon l'information disponible, i.e.
+les données brutes décrivant le système.
+Dans une approche basée sur les données, celles-ci sont la base de notre cadre et la robustesse sera déterminée par leur structure.
+\item A partir des données sont capturés des indicateurs (fonctions objectifs).
+Nous supposons qu'un choix d'indicateurs est une intention particulière de traduire des aspects particuliers du système, i.e.
+de capturer une réalisation d'un ``fait urbain'' au sens de \noun{Mangin}~\cite{mangin1999projet} - une sorte de fait stylisé en terme de processus et de mécanismes, ayant différentes réalisations sur des systèmes distincts dans l'espace, dépendant de chaque contexte géographique précis.
+\item Etant donné plusieurs systèmes et indicateurs associés, un espace commun peut être construit pour les comparer.
+Dans cet espace, les données représentent plus ou moins bien le système réel, c'est à dire qu'elles sont imprécises en fonction de l'échelle initiale, de la précision effective des données.
+Nous proposons de capturer exactement ces différents aspects au travers de la notion de discrépance d'un nuage de points, qui est un outil mathématique provenant des théories d'échantillonnage, permettant d'exprimer la façon dont un jeu de données rempli l'espace dans lequel il s'insère~\cite{dick2010digital}.
 \end{itemize}
 
-Synthesizing these requirements, we propose a notion of \emph{Robustness} of an evaluation that captures both, by combining data reliability with relative importance,
+Synthétisant ces contraintes, nous proposons une notion de \emph{Robustesse} d'une évaluation qui capture à la fois, en combinant la fiabilité des données à l'importance relative des indicateurs,
+
 \begin{enumerate}
-\item \emph{Missing Data} : an evaluation based on more refined datasets will naturally be more robust.
-\item \emph{Indicator importance} : indicators with more relative influence will weight more on the total robustness.
+\item \emph{Données manquantes} : une évaluation se basant sur des jeux de données plus raffinés sera naturellement plus robuste.
+\item \emph{Importance des indicateurs} : les indicateurs avec plus d'importance relative pèseront plus dans la robustesse totale.
 \end{enumerate}
@@
 %%%%%%%%%%%%%%%%
-\subsection{Formal Description}
+\subsection{Description Formelle}
 
 %%%%%%%%%%%%%%%%
-\paragraph{Indicators}
+\paragraph{Indicateurs}
 
-Let $(S_{i})_{1\leq i\leq N}$ be a finite number of geographically disjoints territorial systems, that we assume described through raw data and intermediate indicators, yielding $S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}$ with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$ such that each subspace contain real matrices : $\mathcal{X}_{i,k}=\mathbb{R}^{n_{i,k}^{X}p_{i,k}^{X}}$ (the same holding for $\mathcal{Y}_{i}$).
-We also define an ontological index function $I_{X}(i,k)$ (resp.
-$I_{Y}(i,k)$) taking integer values which coincide if and only if the two variables have the same ontology in the sense of~\cite{livet2010}, i.e.
-they are supposed to represent the same real object.
-We distinguish ``raw data'' $\mathbf{X}_{i}$ from which indicators are computed via explicit deterministic functions, from ``intermediate indicators'' $\mathbf{Y}_{i}$ that are already integrated and can be e.g.
-outputs of elaborated models simulating some aspects of the urban system.
-We define the partial characteristic space of the ``urban fact'' by
+Soit $(S_{i})_{1\leq i\leq N}$ un nombre fini de systèmes territoriaux géographiquement disjoints, % TODO : Q pourquoi nécessaire des les avoir spatially disjoints, could be different indicators on the same area ?
+maybe makes less sense ?
+missing point for comparability ?
+que nous supposons décrits par les données brutes et des indicateurs intermédiaires, donnés par $S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}$ avec $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$ tel que chaque sous-espace contient des matrices réelles : $\mathcal{X}_{i,k}=\mathbb{R}^{n_{i,k}^{X}p_{i,k}^{X}}$ (de la même façon pour $\mathcal{Y}_{i}$).
+Nous définissons également une fonction d'indice ontologique $I_{X}(i,k)$ (resp.
+$I_{Y}(i,k)$) prenant des valeurs entières qui coincident si et seulement si les deux variables ont même ontologie au sens de~\cite{livet2010}, c'est à dire qu'elles sont supposées représenter le même objet réel.
+On distingue les ``données brutes'' $\mathbf{X}_{i}$ à partir desquelles les indicateurs sont calculés généralement par des fonctions déterministes explicites, % TODO not that free on the computation here !
+ des ``indicateurs intermédiaires'' $\mathbf{Y}_{i}$ qui sont déjà intégrés et peuvent être par exemple les sorties de modèles élaborés simulant certains aspects du système urbain.
+Nous définissons l'espace caractéristique du ``fait urbain'' par
 
 \begin{equation}
-%\begin{split}
-%(\mathcal{X},\mathcal{Y}) & \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right)\\
-%& =  \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
-%\end{split}
 (\mathcal{X},\mathcal{Y}) \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right) = \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
@@
 
-with $\mathcal{D}_{\mathcal{X}}=\{\mathcal{X}_{i,k}|I(i,k)\textrm{ distincts},n_{i,k}^{X}\mbox{ maximal}\}$
-(the same holding for $\mathcal{Y}_{i}$).
-It is indeed the abstract space on which indicators are integrated.
-The indices $c$ introduced as a definition here correspond to different indicators across all systems.
-This space is the minimal space common to all systems allowing a common definition for indicators on each.
+avec $\mathcal{D}_{\mathcal{X}}=\{\mathcal{X}_{i,k}|I(i,k)\textrm{ distincts},n_{i,k}^{X}\mbox{ maximal}\}$
+(de même pour $\mathcal{Y}_{i}$).
+Il s'agit en fait de l'espace abstrait sur lequel les indicateurs sont intégrés.
+Les indices $c$ introduit par définition correspondent aux différents indicateurs au sein des différents systèmes.
+Cette espace est l'espace minimal commun à tous les systèmes permettant une définition commune des indicateurs pour tous.
 
-Let $\mathbf{X}_{i,c}$ be the data canonically projected in the corresponding subspace, well defined for all $i$ and all $c$.
-We make the key assumption that all indicators are computed by integration against a certain kernel, i.e.
-that for all $c$, there exists $H_{c}$ space of real-valued functions on $(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})$, such that for all $h\in H_{c}$ :
+Soit $\mathbf{X}_{i,c}$ les données projetées canoniquement sur le sous-espace correspondant, bien définies pour tout $i$ et tout $c$.
+Nous faisons donc l'hypothèse clé que tous les indicateurs sont calculés par intégration contre un noyau donné, i.e.
+pour tout $c$ il existe $H_{c}$ espace de fonctions à valeurs réelles sur $(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})$, tel que pour tout $h\in H_{c}$ :
+
 \begin{enumerate}
-\item $h$ is ``enough'' regular (tempered distributions e.g.)
-\item $q_c=\int_{(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})}h$ is a function describing the ``urban fact'' (the indicator in itself)
+\item $h$ est ``suffisamment'' régulière (distribution tempérée par exemple)
+\item $q_c=\int_{(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})}h$ est une fonction décrivant le ``fait urbain'' (l'indicateur en lui-même)
 \end{enumerate}
 
-Typical concrete example of kernels can be :
+Des exemples typiques de noyaux peuvent être :
 
 \begin{itemize}
-\item A mean of rows of $\mathbf{X}_{i,c}$ is computed with $h(x)=x\cdot f_{i,c}(x)$ where $f_{i,c}$ is the density of the distribution of the assumed underlying variable.
-\item A rate of elements respecting a given condition $C$, $h(x)=f_{i,c}(x)\chi_{C(x)}$ 
-\item For already aggregated variables $\mathbf{Y}$, a Dirac distribution allows to express them also as a kernel integral. 
+\item Une moyenne des lignes de $\mathbf{X}_{i,c}$ est calculée par $h(x)=x\cdot f_{i,c}(x)$ où $f_{i,c}$ est la densité de la distribution de la variable sous-jacente.
+\item Un taux d'éléments du jeu de données respectant une condition donnée $C$, $h(x)=f_{i,c}(x)\chi_{C(x)}$.
+\item Pour des variables déjà agrégées $\mathbf{Y}$, une distribution de Dirac permet des les exprimer également comme des intégrales de noyaux.
 \end{itemize}
@@
 %%%%%%%%%%%%%%%%
-\paragraph{Aggregation}
+\paragraph{Agrégation}
 
-%As described in introduction, the multi-objective nature of the problem is tackled through linear aggregation.
-Although it may be seen as a reductive approach, it becomes an asset for the application of integral approximation theorems as the next step of our method.
-Weighting objectives in multi-attribute decision-making is indeed the crucial point of the processes, and numerous methods are available (see~\cite{wang2009review} for a review for the particular case of sustainable energy management).
-Let define weights for the linear aggregation.
-We assume the indicators normalized, i.e.
-$q_c \in [0,1]$, for a more simple construction of relative weights.
-For $i,c$ and $h_{c}\in H_{c}$ given, the weight $w_{i,c}$ is simply constituted by the relative importance of the indicator $w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}$ where $\hat{q}_{i,c}$ is an estimator of $q_{c}$ for data $\mathbf{X}_{i,c}$ (i.e.
-the effectively calculated value).
-Note that this step can be extended to any sets of weight attributions, by taking for example $\tilde{w}_{i,c} = w_{i,c} \cdot w'_{i,c}$ if $\mathbf{w}'$ are the weights attributed by the decision-maker.
-We focus here on the relative influence of attributes and thus choose this simple form for weights.
-
-%  subjective part is shit.
-% subjective weight : various potential methods (notations, importance order, etc.), reviewed in~\cite{wang2009review}, are available to give a normalized subjective weight.
-In practice, these will be set to zero in our first implementation as we test the framework on virtual cases.
+La détermination des poids est en fait le point crucial des processus de prise de décision multi-attributs, et de nombreuses méthodes sont disponibles (voir~\cite{wang2009review} pour une revue dans le cas particulier de la gestion de l'énergie durable).
+Définissons les poids pour l'agrégation linéaire.
+Nous supposons les indicateurs normalisés, i.e.
+$q_c \in [0,1]$, pour une construction plus simple des poids relatifs.
+% TODO : indeed $h_c \in [0,1]$ is the right assumption
+Pour $i,c$ et $h_{c}\in H_{c}$ donnés, le poids $w_{i,c}$ est simplement constitué par l'importance relative de l'indicateur $w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}$ où $\hat{q}_{i,c}$ est un estimateur de $q_{c}$ pour les données $\mathbf{X}_{i,c}$ (i.e.
+la valeur calculée effectivement).
+On peut noter que cette étape n'est pas contraignante et que cela peut être étendu à tout ensemble d'attribution de poids, en prenant par exemple $\tilde{w}_{i,c} = w_{i,c} \cdot w'_{i,c}$ si $\mathbf{w}'$ sont les poids fixés par le preneur de décisions.
+Nous nous concentrons sur l'influence relative des attributs et pour cela choisissons cette forme simple pour les poids.
 
 %%%%%%%%%%%%%%%%
-\paragraph{Robustness Estimation}
+\paragraph{Estimation de la Robustesse}
 
-The scene is now set up to be able to estimate the robustness of the evaluation done through the aggregated function.
-Therefore, we apply an integral approximation method similar to methods introduced in~\cite{varet2010developpement}, since the integrated form of indicators indeed brings the benefits of such powerful theoretical results.
-Let $\mathbf{X}_{i,c}=(\vec{X}_{i,c,l})_{1\leq l\leq n_{i,c}}$ and $D_{i,c}=Disc_{\tilde{\mathcal{X}}_{c},L^2}(\mathbf{X}_{i,c})$ the discrepancy of data points cloud\footnote{The discrepancy is defined as the $L2$-norm of local discrepancy which is for normalized data points $\mathbf{X}=(x_{ij})\in \left[0,1\right]^d$, a function of $\mathbf{t}\in \left[0,1\right]^d$ comparing the number of points falling in the corresponding hypercube with its volume, by $disc(\mathbf{t}) = \frac{1}{n}\sum_i \mathbbm{1}_{\prod_j x_{ij}<t_j} - \prod_j t_j$.
-It is a measure of how the point cloud covers the space.}~\cite{niederreiter1972discrepancy}.
-With $h\in H_{c}$, we have the upper bound on the integral approximation error
+La scène est à présent prête pour permettre d'estimer la robustesse d'une évaluation faite par la fonction d'agrégation.
+Pour cela, nous appliquons un théorème d'approximation d'intégrale similaire au méthodes introduites dans~\cite{varet2010developpement}, puisque la forme intégrée des indicateurs permet justement de bénéficier de tels résultats théoriquement puissant.
+Soit $\mathbf{X}_{i,c}=(\vec{X}_{i,c,l})_{1\leq l\leq n_{i,c}}$ et $D_{i,c}=Disc_{\tilde{\mathcal{X}}_{c},L^2}(\mathbf{X}_{i,c})$ le discrépance du jeu de données\footnote{La discrépance est définie comme la norme-$L2$ de la discrépance locale qui est pour des points de données normalisés $\mathbf{X}=(x_{ij})\in \left[0,1\right]^d$, une fonction de $\mathbf{t}\in \left[0,1\right]^d$ comparant le nombre de points compris dans le volume de l'hypercube correspondant, donné par $disc(\mathbf{t}) = \frac{1}{n}\sum_i \mathbbm{1}_{\prod_j x_{ij}<t_j} - \prod_j t_j$.
+C'est une mesure de la manière dont le nuage de points couvre l'espace.}~\cite{niederreiter1972discrepancy}.
+Avec $h\in H_{c}$, on a la borne supérieure sur l'erreur d'approximation de l'intégrale
 
@@
 
-where $K$ is a constant independent of data points and objective function. It directly yields
+où $K$ est une constante indépendante des points de données et des fonctions objectifs. Cela donne directement
 
@@
 
-Assuming the error reasonably realized (``worst case'' scenario for knowledge of the theoretical value of aggregated function), we take this upper bound as an approximation of its magnitude.
-Furthermore, taking normalized indicators implies $\left|\left|\left|h_c\right|\right|\right| = 1$.
-We propose then to compare error bounds between two evaluations.
-They depend only on data distribution (equivalent to \emph{statistical robustness}) and on indicators chosen (sort of \emph{ontological robustness}, i.e.
-do the indicators have a real sense in the chosen context and do their values make sense), and are a way to combine these two type of robustnesses into a single value.
+En supposant l'erreur réalisée de manière raisonnable (scénario du ``pire de cas'' pour la connaissance de la valeur théorique de la fonction agrégée), nous prenons cette borne supérieure comme une approximation de sa magnitude.
+De plus, la normalisation des indicateurs implique que $\left|\left|\left|h_c\right|\right|\right| = 1$.
+Nous proposons alors de comparer les bornes d'erreurs entre deux évaluations.
+Elle dépendent seulement de la distribution des données (équivalence à la \emph{robustesse statistique}) et des indicateurs choisis (sorte de \emph{robustesse ontologique}, i.e.
+est-ce que les indicateurs ont un sens réel dans le contexte choisi et est-ce que leur valeur fait sens), et sont un moyen de combiner ces deux types de robustesse dans une seule valeur.
 
-We thus define a \emph{robustness ratio} to compare the robustness of two evaluations by
+Nous définissons ainsi un \emph{ratio de robustesse} pour comparer la robustesse de deux évaluations par
 
@@
 
-The intuitive sense of this definition is that one compares robustness of evaluations by comparing the highest error done in each based on data structure and relative importance.
+L'interprétation intuitive de cette définition est que l'on compare la robustesse des évaluations en comparant la plus grande erreur faite dans chaque cas selon la structure des données et l'importance relative.
 
-By taking then an order relation on evaluations by comparing the position of the ratio to one, it is obvious that we obtain a complete order on all possible evaluations.
-This ratio should theoretically allow to compare any evaluation of an urban system.
-To keep an ontological sense to it, it should be used to compare disjoints sub-systems with a reasonable proportion of indicators in common, or the same sub-system with varying indicators.
-Note that it provides a way to test the influence of indicators on an evaluation by analyzing the sensitivity if the ratio to their removal.
-On the contrary, finding a ``minimal'' number of indicators each making the ratio strongly vary should be a way to isolate essential parameters ruling the sub-system.
+En construisant une relation d'ordre sur les évaluations en comparant la position du ratio par rapport à un, il est clair qu'on obtient un ordre complet sur l'ensemble des évaluations possibles.
+Ce ratio devrait en théorie permettre de comparer n'importe quelle évaluation d'un système urbain.
+Afin de garder un sens ontologique à cela, il devrait être utilisé pour comparer des sous-systèmes disjoints  avec une proportion raisonnable d'indicateurs en commun, ou le même sous-système avec des indicateurs différents.
+On peut noter que cela fournit un moyen de tester l'influence des indicateurs sur une évaluation, en analysant la sensibilité du ratio à leur suppression.
+Au contraire, la détermination d'un nombre ``minimal'' d'indicateurs faisant chacun varier le ratio fortement pourrait être un moyen d'isoler des paramètres essentiels régissant le sous-système.
 
@@
 %%%%%%%%%%%%%%%%
-\section{Results}
-
-%We apply our framework to a ``toy-model'', as elaborated indicator definition nor raw data collection was not the primary aim of this work, which was more to theoretically introduce the framework.
-It is however essential for such a data-driven paradigm to demonstrate through simulation the potentialities of the method, what we do here on semi-synthetic data.
+\section{Résultats}
 
 %%%%%%%%%%%%%%%%
-\paragraph{Implementation}
+\paragraph{Implémentation}
 
-Preprocessing of geographical data is made through QGIS~\cite{qgis2011quantum} for performance reasons.
-Core implementation of the framework is done in R~\cite{team2000r} for the flexibility of data management and statistical computations.
-Furthermore, the package \texttt{DiceDesign}~\cite{franco20092} written for numerical experiments and sampling purposes, allows an efficient and direct computation of discrepancies.
-Last but not least, all source code is openly available on the \texttt{git} repository of the project\footnote{at \url{https://github.com/JusteRaimbault/RobustnessDiscrepancy}} for reproducibility purposes~\cite{ram2013git}.
+Le pré-traitement des données géographiques est fait via QGIS~\cite{qgis2011quantum} pour des raisons de performances. % TODO plutôt ergonomie ?
+L'implémentation du coeur est faite en R~\cite{team2000r} pour la flexibilité de la gestion des données et du traitement statistique.
+De plus, le package \texttt{DiceDesign}~\cite{franco20092} conçu pour les expériences numériques et l'échantillonnage, permet un calcul efficient et direct des discrépances.
+Enfin, tout aussi important, l'ensemble du code source est disponible de manière ouverte sur le dépôt \texttt{git}du projet\footnote{à \url{https://github.com/JusteRaimbault/RobustnessDiscrepancy}} pour permettre la reproductibilité et la réutilisation~\cite{ram2013git}.
 
 %%%%%%%%%%%%%%%%
-\subsection{Implementation on Synthetic Data}
+\subsection{Implémentation sur des Données Synthétiques}
 
-We propose in a first time to illustrate the implementation with an application to synthetic data and indicators, for intra-urban quality indicators in the city of Paris.
+Nous proposons dans un premier temps d'illustrer l'implémentation par une application à des données et indicateurs synthétiques, pour des indicateurs de qualité de vie intra-urbaine pour la ville de Paris.
 
 %%%%%%%%%%%%%%%%
-\paragraph{Data Collection}
+\paragraph{Données Brutes}
 
-We base our virtual case on real geographical data, in particular for \emph{arrondissements} of Paris.
-We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features.
-We use the street network and position of buildings within the city of Paris.
-%\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}
-Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source.
-We use centroids of buildings polygons, and segments of street network.
-Dataset overall consists of around $200k$ building features and $100k$ road segments.
-%An example of visualization of used data is shown in Fig.~\ref{fig:data_ex}.
+Le cas virtuel se base sur des données géographiques réelle, en particulier pour les arrondissements parisiens.
+Nous utilisons les données disponibles par le projet OpenStreetMap~\cite{bennett2010openstreetmap} qui fournit déjà des données précises en haute définition pour de nombreux aspects urbains.
+Nous utilisons le réseau de rues et la position des bâtiments dans la ville de Paris.
+Les limites des arrondissements, utilisées pour agréger et extraire les features lorsqu'on travaille sur un seul district, sont aussi pris de la même source.
+Nous utilisons les centroïdes des polygones des bâtiments et les segments du réseau de rues.
+Le jeu de données brutes consiste d'environ $200k$ bâtiments et $100k$ segments de rues.
 
 %%%%%%%%%%%%%%%%
-\paragraph{Virtual Cases}
+\paragraph{Indicateurs synthétiques}
 
-We work on each district of Paris (from the 1st to the 20th) as an evaluated urban system.
-We construct random synthetic data associated to spatial features, so each district has to be evaluated many time to obtain mean statistical behavior of toy indicators and robustness ratios.
-The indicators chosen need to be computed on residential and street network spatial data.
-We implement two mean kernels and a conditional mean to show different examples, linked to environmental sustainability and quality of life, that are required to be maximized.
-Note that these indicators have a real meaning but no particular reason to be aggregated, they are chosen here for the convenience of the toy model and the generation of synthetic data.
-With $a\in \{1\ldots 20\}$ the number of the district, $A(a)$ corresponding spatial extent, $b\in B$ building coordinates and $s\in S$ street segments, we take
+Nous travaillons sur chaque arrondissement de Paris (du 1er au 20ème) comme un système urbain évalué.
+Des données synthétiques aléatoires sont associées aux features spatiales, chaque arrondissement pouvant alors être évalué de manière stochastique, et des répétitions permettent d'obtenir le comportement statistique moyen des indicateurs jouets et des ratios de robustesse.
+Les indicateurs choisis doivent être calculés comme des indicateurs résidentiels et du réseau de rues.
+Pour montrer différents exemples, nous implémentons deux kernels moyens et une moyenne conditionnelle, tous liés à la durabilité environnementale et la qualité de vie, chacun devant être maximisés.
+On peut noter que ces indicateurs ont un sens réel mais pas de raison particulière d'être agrégés, ils sont ici choisis pour l'aspect pratique du modèle jouet et de la génération de données synthétiques.
+Avec $a\in \{1\ldots 20\}$ le nombre d'arrondissements, $A(a)$ l'aire spatiale correspondante à chacun, $b\in B$ les coordonnées des bâtiments et $s\in S$ les segments de rues, nous prenons
+
 \begin{itemize}
-\item Complementary of the average daily distance to work with car per individual, approximated by, with $n_{cars}(b)$ number of cars in the building (randomly generated by associated of cars to a number of building proportional to motorization rate $\alpha_m ~ 0.4$ in Paris), $d_w$ distance to work of individuals (generated from the building to a uniformly generated random point in spatial extent of the dataset), and $d_{max}$ the diameter of Paris area, $\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}$
-\item Complementary of average car flows within the streets in the district, approximated by, with $\varphi(s)$ relative flow in street segment $s$, generated through the minimum of 1 and a log-normal distribution adjusted to have $95\%$ of mass smaller than 1 what mimics the hierarchical distribution of street use (corresponding to betweenness centrality), and $l(s)$ segment length, $\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}$
-
-\item Relative length of pedestrian streets $\bar{p}$, computed through a randomly uniformly generated dummy variable adjusted to have a fixed global proportion of segments that are pedestrian.
+\item Le complémentaire de la distance journalière moyenne au travail en voiture par individu, approché par, avec $n_{cars}(b)$ nombre de voiture dans le bâtiment (généré aléatoirement en associant des voiture à bâtiments proportionnel au taux de motorisation attendu $\alpha_m ~ 0.4$ à Paris), $d_w$ distance des individus à leur travail (généré à partir du bâtiment vers un point aléatoire distribué uniformément dans l'étendue spatiale du jeu de données), et $d_{max}$ le diamètre de l'aire de Paris, $\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}$
+\item Le complémentaire des flots moyens de voitures des rues dans la zone, approché par, avec $\varphi(s)$ flot relatif dans le segment de rue $s$, généré par le minimum entre 1 et une distribution log-normale ajustée pour avoir $95\%$ de masse plus petite que 1, ce qui mimique la distribution hiérarchique de l'utilisation des rues (qui correspond à la centralité de chemin), et $l(s)$ longueur du segment, $\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}$
+\item Longueur relative de rues piétonnes $\bar{p}$, calculé vie une dummy variable aléatoire uniforme ajustée pour obtenir une proportion fixée de segments pédestre.
 \end{itemize}
@@
 \hline
-1 th & 0.731655 $\pm$ 0.041099 & 0.917462 $\pm$ 0.026637 & 0.191615 $\pm$ 0.052142 & 1.000000 $\pm$ 0.000000\\[3pt]
+1er & 0.731655 $\pm$ 0.041099 & 0.917462 $\pm$ 0.026637 & 0.191615 $\pm$ 0.052142 & 1.000000 $\pm$ 0.000000\\[3pt]
 \hline
-2 th & 0.723225 $\pm$ 0.032539 & 0.844350 $\pm$ 0.036085 & 0.209467 $\pm$ 0.058675 & 1.002098 $\pm$ 0.039972\\[3pt]
+2ème & 0.723225 $\pm$ 0.032539 & 0.844350 $\pm$ 0.036085 & 0.209467 $\pm$ 0.058675 & 1.002098 $\pm$ 0.039972\\[3pt]
 \hline
-3 th & 0.713716 $\pm$ 0.044789 & 0.797313 $\pm$ 0.057480 & 0.185541 $\pm$ 0.065089 & 0.999341 $\pm$ 0.048825\\[3pt]
+3ème & 0.713716 $\pm$ 0.044789 & 0.797313 $\pm$ 0.057480 & 0.185541 $\pm$ 0.065089 & 0.999341 $\pm$ 0.048825\\[3pt]
 \hline
-4 th & 0.712394 $\pm$ 0.042897 & 0.861635 $\pm$ 0.030859 & 0.201236 $\pm$ 0.044395 & 0.973045 $\pm$ 0.036993\\[3pt]
+4ème & 0.712394 $\pm$ 0.042897 & 0.861635 $\pm$ 0.030859 & 0.201236 $\pm$ 0.044395 & 0.973045 $\pm$ 0.036993\\[3pt]
 \hline
-5 th & 0.715557 $\pm$ 0.026328 & 0.894675 $\pm$ 0.020730 & 0.209965 $\pm$ 0.050093 & 0.963466 $\pm$ 0.040722\\[3pt]
+5ème & 0.715557 $\pm$ 0.026328 & 0.894675 $\pm$ 0.020730 & 0.209965 $\pm$ 0.050093 & 0.963466 $\pm$ 0.040722\\[3pt]
 \hline
-6 th & 0.733249 $\pm$ 0.026890 & 0.875613 $\pm$ 0.029169 & 0.206690 $\pm$ 0.054850 & 0.990676 $\pm$ 0.031666\\[3pt]
+6ème & 0.733249 $\pm$ 0.026890 & 0.875613 $\pm$ 0.029169 & 0.206690 $\pm$ 0.054850 & 0.990676 $\pm$ 0.031666\\[3pt]
 \hline
-7 th & 0.719775 $\pm$ 0.029072 & 0.891861 $\pm$ 0.026695 & 0.209265 $\pm$ 0.041337 & 0.966103 $\pm$ 0.037132\\[3pt]
+7ème & 0.719775 $\pm$ 0.029072 & 0.891861 $\pm$ 0.026695 & 0.209265 $\pm$ 0.041337 & 0.966103 $\pm$ 0.037132\\[3pt]
 \hline
-8 th & 0.713602 $\pm$ 0.034423 & 0.931776 $\pm$ 0.015356 & 0.208923 $\pm$ 0.036814 & 0.973975 $\pm$ 0.033809\\[3pt]
+8ème & 0.713602 $\pm$ 0.034423 & 0.931776 $\pm$ 0.015356 & 0.208923 $\pm$ 0.036814 & 0.973975 $\pm$ 0.033809\\[3pt]
 \hline
-9 th & 0.712441 $\pm$ 0.027587 & 0.910817 $\pm$ 0.015915 & 0.202283 $\pm$ 0.049044 & 0.971889 $\pm$ 0.035381\\[3pt]
+9ème & 0.712441 $\pm$ 0.027587 & 0.910817 $\pm$ 0.015915 & 0.202283 $\pm$ 0.049044 & 0.971889 $\pm$ 0.035381\\[3pt]
 \hline
-10 th & 0.713072 $\pm$ 0.028918 & 0.881710 $\pm$ 0.021668 & 0.210118 $\pm$ 0.040435 & 0.991036 $\pm$ 0.038942\\[3pt]
+10ème & 0.713072 $\pm$ 0.028918 & 0.881710 $\pm$ 0.021668 & 0.210118 $\pm$ 0.040435 & 0.991036 $\pm$ 0.038942\\[3pt]
 \hline
-11 th & 0.682905 $\pm$ 0.034225 & 0.875217 $\pm$ 0.019678 & 0.203195 $\pm$ 0.047049 & 0.949828 $\pm$ 0.035122\\[3pt]
+11ème & 0.682905 $\pm$ 0.034225 & 0.875217 $\pm$ 0.019678 & 0.203195 $\pm$ 0.047049 & 0.949828 $\pm$ 0.035122\\[3pt]
 \hline
-12 th & 0.646328 $\pm$ 0.039668 & 0.920086 $\pm$ 0.019238 & 0.198986 $\pm$ 0.023012 & 0.960192 $\pm$ 0.034854\\[3pt]
+12ème & 0.646328 $\pm$ 0.039668 & 0.920086 $\pm$ 0.019238 & 0.198986 $\pm$ 0.023012 & 0.960192 $\pm$ 0.034854\\[3pt]
 \hline
-13 th & 0.697512 $\pm$ 0.025461 & 0.890253 $\pm$ 0.022778 & 0.201406 $\pm$ 0.030348 & 0.960534 $\pm$ 0.033730\\[3pt]
+13ème & 0.697512 $\pm$ 0.025461 & 0.890253 $\pm$ 0.022778 & 0.201406 $\pm$ 0.030348 & 0.960534 $\pm$ 0.033730\\[3pt]
 \hline
-14 th & 0.703224 $\pm$ 0.019900 & 0.902898 $\pm$ 0.019830 & 0.205575 $\pm$ 0.038635 & 0.932755 $\pm$ 0.033616\\[3pt]
+14ème & 0.703224 $\pm$ 0.019900 & 0.902898 $\pm$ 0.019830 & 0.205575 $\pm$ 0.038635 & 0.932755 $\pm$ 0.033616\\[3pt]
 \hline
-15 th & 0.692050 $\pm$ 0.027536 & 0.891654 $\pm$ 0.018239 & 0.200860 $\pm$ 0.024085 & 0.929006 $\pm$ 0.031675\\[3pt]
+15ème & 0.692050 $\pm$ 0.027536 & 0.891654 $\pm$ 0.018239 & 0.200860 $\pm$ 0.024085 & 0.929006 $\pm$ 0.031675\\[3pt]
 \hline
-16 th & 0.654609 $\pm$ 0.028141 & 0.928181 $\pm$ 0.013477 & 0.202355 $\pm$ 0.017180 & 0.963143 $\pm$ 0.033232\\[3pt]
+16ème & 0.654609 $\pm$ 0.028141 & 0.928181 $\pm$ 0.013477 & 0.202355 $\pm$ 0.017180 & 0.963143 $\pm$ 0.033232\\[3pt]
 \hline
-17 th & 0.683020 $\pm$ 0.025644 & 0.890392 $\pm$ 0.023586 & 0.198464 $\pm$ 0.033714 & 0.941025 $\pm$ 0.034951\\[3pt]
+17ème & 0.683020 $\pm$ 0.025644 & 0.890392 $\pm$ 0.023586 & 0.198464 $\pm$ 0.033714 & 0.941025 $\pm$ 0.034951\\[3pt]
 \hline
-18 th & 0.699170 $\pm$ 0.025487 & 0.911382 $\pm$ 0.027290 & 0.188802 $\pm$ 0.036537 & 0.950874 $\pm$ 0.028669\\[3pt]
+18ème & 0.699170 $\pm$ 0.025487 & 0.911382 $\pm$ 0.027290 & 0.188802 $\pm$ 0.036537 & 0.950874 $\pm$ 0.028669\\[3pt]
 \hline
-19 th & 0.655108 $\pm$ 0.031857 & 0.884214 $\pm$ 0.027816 & 0.209234 $\pm$ 0.032466 & 0.962966 $\pm$ 0.034187\\[3pt]
+19ème & 0.655108 $\pm$ 0.031857 & 0.884214 $\pm$ 0.027816 & 0.209234 $\pm$ 0.032466 & 0.962966 $\pm$ 0.034187\\[3pt]
 \hline
-20 th & 0.637446 $\pm$ 0.032562 & 0.873755 $\pm$ 0.036792 & 0.196807 $\pm$ 0.026001 & 0.952410 $\pm$ 0.038702\\[3pt]
+20ème & 0.637446 $\pm$ 0.032562 & 0.873755 $\pm$ 0.036792 & 0.196807 $\pm$ 0.026001 & 0.952410 $\pm$ 0.038702\\[3pt]
 \hline
@@
 
-\caption{Numerical results of simulation for each district with $N=50$ repetitions.
-Each toy indicator value is given by mean on repetitions and associated standard deviation.
-Robustness ratio is computed relative to first district (arbitrary choice).
-A ratio smaller than 1 means that integral bound is smaller for upper district, i.e.
-that evaluation is more robust for this district.
-Because of the small size of first district, we expected a majority of district to give ratio smaller than 1, what is confirmed by results, even when adding standard deviations.}
+\caption{Résultats numériques des simulations pour chaque arrondissement avec $N=50$ répétitions.
+Chaque valeur des indicateurs factice est donnée par sa moyenne sur les répétitions et la déviation standard associée.
+Le ratio de robustesse est calculé par rapport au premier arrondissement (choix arbitraire).
+Un ratio inférieur à 1 signifie que la borne de l'intégrale est plus petite pour le premier système, i.e.
+que l'évaluation est plus robuste pour celui-ci.
+% TODO : wrong à l'oral 15th block size ?
+ A cause de la petite taille du 1er arrondissement, on s'attend que la majorité des arrondissements aient un ratio plus petit que 1, ce qui se confirme dans les résultats même lorsque l'on ajoute la déviation standard.}
 
@@
 
-As synthetic data are stochastic, we run the computation for each district $N=50$ times, what was a reasonable compromise between statistical convergence and time required for computation.
-Table 1 shows results (mean and standard deviations) of indicator values and robustness ratio computation.
-Obtained standard deviation confirm that this number of repetitions give consistent results.
-Indicators obtained through a fixed ratio show small variability what may a limit of this toy approach.
-However, we obtain the interesting result that a majority of districts give more robust evaluations than 1st district, what was expected because of the size and content of this district : it is indeed a small one with large administrative buildings, what means less spatial elements and thus a less robust evaluation following our definition of the robustness.
+Comme les données synthétiques sont stochastiques, les simulations sont lancées pour chaque quartier $N=50$ fois, ce qui était un compromis raisonnable entre convergence statistique et temps nécessaire au calcul.
+La table 1 montre les résultats (moyennes et déviations standard) des valeurs des indicateurs et le calcul du ratio de robustesse.
+Les déviations standard obtenues confirment que ce nombre de simulations donnent des résultats consistants.
+Les indicateurs obtenus en fixant un ratio fixe montre peu de variabilité, ce qui peut être une limite de cette approche jouet.
+On obtient toutefois le résultat intéressant que la majorité des arrondissements donne des évaluations plus robustes que le 1er arrondissement, ce qui était attendu par la taille et la fonction de ce quartier: il s'agit en effet d'un petit quartier avec de grand bâtiment administratifs, ce qui implique moins d'éléments spatiaux et pour cela une évaluation moins robuste selon la définition qu'on en a donnée.
+
+% TODO : question : check the claim on block size, maybe it's only because bigger ?
+%   -- -> cf up --
 
 %%%%%%%%%%%%%%%%
-\subsection{Application to a Real Case : Metropolitan Segregation}
+\subsection{Application à un Cas Réel: Ségrégation Métropolitaine}
 
-The first example was aimed to show potentialities of the method but was purely synthetic, hence yielding no concrete conclusion nor implications for policy.
-We propose now to apply it to real data for the example of metropolitan segregation.
+Le premier exemple avait pour but de montrer les potentialités de la méthode mais était purement synthétique, ne pouvant pour cela fournir pas de conclusion concrete ni d'implications pour la gouvernance.
+Nous proposons maintenant de l'appliquer à des données réelles dans le cas de la ségrégation métropolitaine.
 
-\paragraph{Data}
+\paragraph{Données}
 
-We work on income data available for France at an intra-urban level (basic statistical units IRIS) for the year 2011 under the form of summary statistics (deciles if the area is populated enough to ensure anonymity), provided by INSEE\footnote{\texttt{http://www.insee.fr}}.
-Data are associated with geographical extent of statistical units, allowing computation of spatial analysis indicators.
+Nous travaillons sur les données de revenus, disponible pour la France à un niveau intra-urbain (unités statistiques élémentaires IRIS) pour l'année 2011 sous la forme de résumé statistiques (déciles uniquement si la zone est peuplée suffisamment pour assurer l'anonymat), fournies par l'INSEE\footnote{\texttt{http://www.insee.fr}}.
+Les données sont associées à l'étendue géographique des unités statistiques, permettant le calcul d'indicateurs d'analyse spatiale.
 
-\paragraph{Indicators}
+\paragraph{Indicateurs}
 
-We use here three indicators of segregation integrated on a geographical area.
-Let assume the area divided into covering units $\mathcal{S}_i$ for $1\leq i \leq N$ with centroids $(x_i,y_i)$.
-Each unit has characteristics of population $P_i$ and median income $X_i$.
-We define spatial weights used to quantify strength of geographical interactions between units $i,j$, with $d_{ij}$ euclidian distance between centroids : $w_{ij} = \frac{P_i P_j}{\left(\sum_k P_k\right)^2}\cdot \frac{1}{d_{ij}}$ if $i\neq i$ and $w_{ii} = 0$.
-The normalized indicators are the following
+Nous utilisons ici trois indicateurs de ségrégation intégrés sur une zone géographique.
+Supposons la zone divisée en unités couvrantes $\mathcal{S}_i$ pour $1\leq i \leq N$ avec pour centroïdes $(x_i,y_i)$.
+Chaque unité a des caractéristiques de population $P_i$ et de revenu médian $X_i$.
+On définit des poids spatiaux utilisés pour quantifier l'intensité des interactions géographiques entre unités $i,j$, avec $d_{ij}$ distance euclidienne entre centroïdes: $w_{ij} = \frac{P_i P_j}{\left(\sum_k P_k\right)^2}\cdot \frac{1}{d_{ij}}$ si $i\neq j$ % TODO typo eng paper
+ et $w_{ii} = 0$. Les indicateurs normalisés sont les suivants
 
 \begin{itemize}
-\item Spatial autocorrelation Moran index, defined as weighted normalized covariance of median income by $\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}$
-\item Dissimilarity index (close to Moran but integrating local dissimilarities rather than correlations), given by $d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|$\\ with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
-\item Complementary of the entropy of income distribution that is a way to capture global inequalities $\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)$
+\item Indice d'autocorrelation spatiale de Moran, défini comme la covariance pondérée normalisée du revenu médian par $\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}$
+\item Indice de dissimilarité (proche du Moran mais intégrant les dissimilarités locales plutôt que les corrélations), donné par $d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|$\\ avec $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
+\item Le complémentaire de l'entropie de la distribution des revenus, qui est une façon de capturer des inégalités globales $\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)$
 \end{itemize}
+
+De nombreuse
 
